---
title: "Reliability, Availability, and Maintainability (RAM)"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(WeibullR.learnr)
```

## Introduction

Welcome to the tutorial on **Reliability, Availability, and Maintainability (RAM)**! In this tutorial, you will learn about these fundamental concepts and their applications.

## Learning Objectives

By the end of this module, learners will be able to:

* Define key reliability metrics, including reliability, availability, and failure rate.
* Describe the significance of MTTR, MTTF, and MTBF in reliability engineering.
* Calculate probability of failure using given reliability data.
* Interpret $B_n$ or $L_n$ life values in the context of product reliability.
* Differentiate between different reliability measures.

## What is RAM?

The terms Reliability, Availability, and Maintainability may sound similar, but they actually represent different concepts.

**Reliability**: The ability of an item to perform its intended function without failure over a specified period.

**Availability**: The proportion of time an item is in a functioning condition.

**Maintainability**: The ease and speed with which an item can be restored to operational status after a failure.

That being said, all 3 terms are inter-related. A car that is not very reliable is probably not available very often. A car that is also difficult to maintain is probably available even less. 

## Reliability

Reliability is the probability that an item will not fail under defined conditions for a specified period.

$$ Reliability = (1 - (Failed Time/Total Time))*100 $$
Where Failed Time is the time that the item was not functioning, and Total Time is the total time that the item was in operation.

Suppose an item ran for 5 years total and was failed for 10 of those days. The reliability of the item is: 

$$ Reliability = (1 - (10/(5*365)))*100 = 94.5% $$
Unreliability is the opposite of reliability, that is, the probability that an item will fail under defined conditions for a specified period. 

$$ Unreliability = 1 - Reliability = (Failed Time/Total Time)*100 $$
Take the previous example, the unreliability of the item is:

$$ Unreliability = 1 - 0.945 = 0.055 = (10/(5*365))*100 = 5.5% $$

```{r quiz1, echo=FALSE}
quiz(caption = "Exercise: Reliability",
  question("An item has run for 3 years total and was failed for 5 of those days. What is the reliability of the item?",
    answer("90%", message = "Not quite, try again."),
    answer("0.5%", message = "0.5% is the unreliability, not the reliability."),
    answer("99.5%", correct = TRUE),
    answer("75%", message = "Not quite, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("What is the unreliability of the item?",
    answer("0.5%", correct = TRUE),
    answer("90%", message = "Not quite, try again."),
    answer("99.5%", message = "99.5% is the reliability, not the unreliability."),
    answer("75%", message = "Not quite, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Availability

Availability is the probability that an item will be available for service. Unavailable time includes failed time and scheduled maintenance.

$$ Availability = (1 - (Unavailable Time/Total Time))*100 $$
Where Unavailable Time is the time that the item was not available for service, and Total Time is the total time that the item was in operation.

Unavailable time is different than standby time, where standby time is when the item is available for service, but is not being used. 

Suppose an item ran for 5 years total, was failed for 10 days, and had scheduled maintenance for 15 days. The availability of the item is:

$$ Availability = (1 - ((10+15)/(5*365)))*100 = 93.2% $$

```{r availability, echo=FALSE}
quiz(caption = "Exercise: Availability",
     question("What is the difference between reliability and availability?",
       answer("Reliability includes scheduled maintenance, availability does not.", message = "Not quite, try again."),
       answer("Availability includes scheduled maintenance, reliability does not.", correct = TRUE),
       answer("There is no difference, they are the same.", message = "There is a slight difference, try again."),
       answer("Reliability is always larger than availability.", message = "Not always, try again."),
       random_answer_order = TRUE,
       allow_retry = TRUE
     ),
  question("An item has run for 3 years total, was failed for 5 days, and had scheduled maintenance for 14 days. What is the availability of the item?",
    answer("95%", message = "Not quite, try again."),
    answer("1.7%", message = "1.7% is too low, try again."),
    answer("98.3%", correct = TRUE),
    answer("50%", message = "50% is too low, try again."),
    random_answer_order = TRUE,
  allow_retry = TRUE
  )
)
```

## Mean Time to Repair

The Mean Time to Repair (MTTR) is a measure of maintainability and is the mean time to repair an item.

$$ MTTR = \sum_{i=1}^n RepairTime_i/RepairCount $$
Where $RepairTime_i$ is the time to repair for the *i*th repair, and $RepairCount$ is the total number of repairs.

Suppose 100 items run for 5 years total. During that time, 5 failures occur with repair times in days of 5, 10, 15, 8, and 12. The MTTR is:

$$ MTTR = (5+10+15+8+12)/5 = 10 days $$

```{r quiz3, echo=FALSE}
quiz(caption = "Exercise: MTTR",
  question("What does MTTR stand for?",
    answer("Mean Time to Recovery", message = "Not quite, try again."),
    answer("Mean Time to Repair", correct = TRUE),
    answer("Mean Time to Reliability", message = "Not quite, try again."),
    answer("Mean Time to Reboot", message = "Not quite, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("1000 items run for 3 years total. During that time, 5 failures occur with repair times in days of 5, 10, 15, 8, and 12. What is the MTTR?",
    answer("15 days", message = "15 days is too high, try again."),
    answer("5 days", message = "5 days is too low, try again."),
    answer("10 days", correct = TRUE),
    answer("8 days", message = "8 days is too low, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Mean Time to Failure

The Mean Time to Failure (MTTF) is a measure of reliability for *non-repairable* items, meaning that once the item fails, it is not repaired, but replaced. The MTTF is the ratio of total time by the number of failures for a given item.

$$ MTTF = Total Time/FailureCount $$
Where *Total Time* is the total time that the item was in operation, and *FailureCount* is the total number of failures.

Suppose 100 items run for 5 years total. During that time, 5 failures occur. The MTTF is:

$$ MTTF = (5*365)/5 = 365 years $$
## Mean Time Between Failures

Mean Time Between Failures (MTBF) is a measure of reliability for *repairable systems*, meaning that once the item fails, it is repaired and put back into service. In this way, the item can fail multiple times. The MTBF is the ratio of total time by the number of failures for a given item.

The calculation is similar to the MTTF, just with a slightly different interpretation. 

Take an example where an item fails 5 times during a total time of 10,000 hours. The MTBF is:

$$ MTBF = 10000/5 = 2000 hours $$

```{r quiz4, echo=FALSE}
quiz(caption = "Exercise: MTTF and MTBF",
  question("What is the difference between MTTF and MTBF?",
    answer("MTTF is for repairable systems, MTBF is for non-repairable systems.", message = "Not quite, try again."),
    answer("MTTF is for non-repairable items, MTBF is for repairable systems.", correct = TRUE),
    answer("There is no difference, they are the same.", message = "There is a slight difference, try again."),
    answer("MTTF is always larger than MTBF.", message = "Not always, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("1000 items run for 3 years total. During that time, 5 failures occur. What is the MTTF?",
    answer("200 years", message = "200 years is too low, try again."),
    answer("0.6 years", message = "0.6 years is too low, try again."),
    answer("600 years", correct = TRUE),
    answer("500 years", message = "500 years is slightly too low, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("An item fails 5 times during a total time of 45,000 hours. What is the MTBF of the item?",
    answer("0.0001 hours", message = "0.0001 hours is the failure rate, not the MTBF."),
    answer("10000 hours", message = "10000 hours is too low, try again."),
    answer("9000 hours", correct = TRUE),
    answer("225,000 hours", message = "225,000 hours is too high, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Failure Rate

The failure rate is the ratio of the number of failures by the total operational time for a given item and a specified period.

A key assumption of MTBF is that the failure rate is constant. Under this assumption, the failure rate is the inverse of the MTBF.

$$ \lambda = 1/MTBF $$
Where $\lambda$ is the failure rate.

The failure rate can also be calculated directly as the number of failures divided by the total operational time.

$$ \lambda = FailureCount/Total Time $$
Where *Total Time* is the total time that the item was in operation, and *FailureCount* is the total number of failures.

Suppose 100 items run for 10 years total. During that time, 20 failures occur. The failure rate is:

$$ \lambda = 20/(10*365) = 0.0055 \text{ failures per year} $$

## Probability of Failure

Failures do not often occur at predictable times and usually occur randomly according to a probability distribution.

When the failure rate is assumed constant, the Exponential distribution is common.

In this case, the cumulative probability of *failure* prior to time *t*, ***F(t)***, is given as the following

$$ F(t) = 1 - e^{-t\lambda} $$
Where $\lambda$ is the failure rate.

The cumulative probability of *survival* prior to time *t*, ***R(t)***, is the opposite of ***F(t)***, that is

$$ R(t) = 1 - F(t) $$
$$ R(t) = e^{-t\lambda} $$

For example, suppose an item has a failure rate of 0.1 failures per year. The probability of survival at time 10 is:

$$ R(10) = e^{-10*0.1} = 0.3679 = 36.79% $$

## $B_n$ or $L_n$ Life

The $B_n$ or $L_n$ life is the time at which *n* % of a population are expected to fail (or 1-*n* % are expected to survive). For example, the B10/L10 life is the time at which 10% of a population are expected to fail, B90/L90 is when 90% are expected to fail, and so on.

To calculate the $B_n$ life, we can solve for *t* in the cumulative probability function. For the exponential function, that is

$$ t = -ln(1-F(t))/\lambda $$
Where $\lambda$ is the failure rate.

For example, suppose an item has a failure rate of 0.2 failures per year. The B10 life is:

$$ B10 = -ln(1-0.1)/0.2 = 0.526 years $$

```{r quiz5, echo=FALSE}
quiz(caption = "Exercise: Failure Rate, Probability of Failure, and Bn Life",
  question("What is the relationship between failure rate and MTBF?",
    answer("Failure rate is the square of MTBF.", message = "Not quite, try again."),
    answer("Failure rate is the inverse of MTBF.", correct = TRUE),
    answer("Failure rate is the same as MTBF.", message = "Not quite, try again."),
    answer("Failure rate is the logarithm of MTBF.", message = "Not quite, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("100 items run for a total time of 5000 hours in a year. Out of these items, 75 have failures. What is the failure rate in hours?",
    answer("0.75", message = "0.75 is too high, try again."),
    answer("6667", message = "6667 is too high, try again."),
    answer("0.00015", correct = TRUE),
    answer("0.015", message = "0.015 is too high, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("An item has a failure probability that follows the exponential distribution with a failure rate of 0.1. What is the probability of survival at time 5?",
    answer("39.3%", message = "39.3% is the failure probability, not the survival probability."),
    answer("60.7%", correct = TRUE),
    answer("6.1%", message = "6.1% is too low, try again."),
    answer("93.9%", message = "93.9% is too high, try again."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("An item has a failure probability that follows the exponential distribution with a failure rate of 0.1. What is the B10?",
    answer("0.105", message = "0.105 is too low, try again."),
    answer("1.05", correct = TRUE),
    answer("60.7%", message = "60.7% is a probability, not a time."),
    answer("39.3%", message = "39.3% is a probability, not a time."),
    random_answer_order = TRUE
  )
)
```

## Summary

Congratulations on completing the Reliability, Availability, and Maintainability (RAM) tutorial!

In this tutorial, we've introduced the concepts of Reliability, Availability, and Maintainability. You have learned to calculate reliability using different assumptions and explored the importance of these concepts in system design.

## References

- Abernethy, R.B. (2004) The New Weibull Handbook. Fifth Edition.

- Aden-Buie G, Schloerke B, Allaire J, Rossell Hayes A (2023). learnr: Interactive Tutorials for R. https://rstudio.github.io/learnr/, https://github.com/rstudio/learnr.

