---
title: "An Introduction to Life Data Analysis"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
```

## Introduction

Welcome to the tutorial on **Life Data Analysis**! In this tutorial, you will learn about the basics of life data analysis and how it applies to various fields, such as engineering, manufacturing, and many others. 

## Learning Objectives

By the end of this module, learners will be able to:

* Describe the purpose of Weibull analysis in reliability engineering.
* Differentiate between types of data censoring, including right-censored and interval-censored data.
* Differentiate between different Weibull models (2-parameter Weibull, 3-parameter Weibull, and Weibayes).
* Apply Median Rank Regression (MRR) and Maximum Likelihood Estimation (MLE) estimation methods to sample datasets.
* Interpret results using plotting methods, including probability plots and contour plots.

## What is Life Data Analysis?

Life data analysis is the study of how systems function over time. Life data helps explain how long a system will last, when it will likely fail, and how often it will need maintenance. For example, in the manufacturing industry, life data analysis can be used to understand the lifespan of machines and equipment, helping to optimize maintenance schedules and reduce downtime.

## Why is Life Data Analysis Important?

Understanding life data analysis helps in predicting the lifespan and maintenance needs of various systems, thereby improving reliability and efficiency. By collecting and modeling life data, we can make predictions about how long machines will last or how long before they will need to be repaired or replaced. 

## Weibull Analysis

### The Weibull Distribution

The Weibull distribution is a continuous probability distribution used in life data analysis and many other fields.

Life data analysis is often referred to as Weibull analysis because the Weibull distribution is commonly used to model life data.

While other distributions can be used to model life data, this tutorial focuses on the Weibull distribution because of its wide application.

### Why the Weibull Distribution?

The Weibull is a **flexible** distribution that can fit many different types of data.

### The Normal Distribution

The normal distribution is another common distribution used in statistics.

Take the following normal distribution with a mean of 5 and standard deviation of 1.5. 

```{r echo=FALSE}
x <- seq(0, 10, by = .1) # Create a sequence of x values
y <- dnorm(x, mean = 5, sd = 1.5) # Calculate the normal density function
plot(x,y, type = "l", col = "blue", xlab = "Variable", ylab = "Probability") # Plot the normal distribution
```

### The Weibull vs Normal

The Weibull can *mimic* the normal and many others. To illustrate, adjust the scale 
parameter in the slider until the Weibull below looks similar to the Normal above.

```{r, echo=FALSE}
sliderInput("eta", "Eta:", min = 1, max = 10, value = 1, step = 0.5) # Slider for eta
plotOutput("distPlot") # Plot output
```

```{r, context="server"}
# Plot the Weibull distribution
output$distPlot <- renderPlot({
  x <- seq(0, 10, by = .1)
  y <- dweibull(x, shape = 3.67, scale = input$eta)
  plot(x,y, type = "l", col = "red", xlab = "Variable", ylab = "Probability")
})
```

<div id="filter-hint">
**Hint:** Set Eta equal to 5 as a starting point.
</div>

### The Cumulative Distribution Function

The cumulative distribution function (CDF) describes the probability of survival 
or failure prior to a certain time *t*. In this case, ***R(t)*** gives the cumulative 
probability of *survival* prior to  time *t* while ***F(t)*** gives the cumulative 
probability of *failure* prior to time *t*.

$$ R(t) = 1 - F(t) = e^{-(t/\eta)^\beta} $$

Where $\beta$ is the shape parameter. In life data analysis $\beta$ has a physical 
interpretation related to the failure rate. The failure rate is the rate at which 
failures occur over time.

-   $\beta$ \< 1 - the failure rate is decreasing
-   $\beta$ = 1 - the failure rate is constant
-   $\beta$ \> 1 - the failure rate is increasing

And $\eta$ is the scale parameter or the quantile at 63.2% probability.

```{r quiz1, echo=FALSE}
quiz(caption = "Quiz: The Weibull Parameters",
  question("Which beta indicates an increasing failure rate?",
    answer("1", message = "Beta of 1 indicates a constant failure rate."),
    answer("0.5", message = "Beta less than 1 indicates a decreasing failure rate."),
    answer("2", correct = TRUE, message = "Correct! Beta greater than 1 indicates an increasing failure rate."),
    answer("0", message = "Beta cannot be 0."),
    random_answer_order = TRUE, 
    allow_retry = TRUE
  ),
  question("What does the parameter eta represent?",
    answer("shape parameter", message = "Eta is the scale parameter, not the shape parameter."),
    answer("scale parameter", correct = TRUE, message = "Correct! Eta is the scale parameter."),
    answer("failure rate", message = "Eta is not the failure rate."),
    answer("mean time between failures", message = "Eta is not the mean time between failures."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("What is the cumulative probability when t = eta?",
    answer("0%", message = "At t = eta, the cumulative probability is not 0%."),
    answer("50%", message = "At t = eta, the cumulative probability is not 50%."),
    answer("63.2%", correct = TRUE, message = "Correct! At t = eta, the cumulative probability is 63.2%."),
    answer("100%", message = "At t = eta, the cumulative probability is not 100%."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("What is the cumulative probability when t = 0?",
    answer("0%", correct = TRUE, message = "Correct! At t = 0, the cumulative probability is 0%."),
    answer("50%", message = "At t = 0, the cumulative probability is not 50%."),
    answer("63.2%", message = "At t = 0, the cumulative probability is not 63.2%."),
    answer("100%", message = "At t = 0, the cumulative probability is not 100%."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## The Weibull Parameters

### Beta
 
Beta is the **shape** parameter of the Weibull distribution. Changing Beta changes 
the **shape** of the cumulative distribution function. To illustrate, use the slider 
below to adjust the shape of the Weibull. 

```{r, echo=FALSE}
sliderInput("beta", "Beta:", min = 0.5, max = 5, value = 1, step = 0.5) # Slider for beta
plotOutput("cumPlot") # Plot output
```

```{r, context="server"}
# Plot the cumulative distribution function
output$cumPlot <- renderPlot({
  x <- seq(0, 50, by = .1)
  plot(pweibull(x, shape = input$beta, scale = 10), col = "blue", type = "l", 
       xlab = "Variable", ylab = "Cumulative Probability") 
})
```

As you adjust Beta, notice how the shape of the Weibull changes. Increasing Beta 
makes the curve steeper, indicating an increasing failure rate. Decreasing Beta 
makes the curve less steep, indicating a decreasing failure rate.

### Eta

Eta is the **scale** parameter of the Weibull distribution. Changing Eta changes 
the **scale** of the cumulative distribution function. To illustrate, use the slider 
below to adjust the scale of the Weibull. 

```{r, echo=FALSE}
sliderInput("eta2", "Eta:", min = 1, max = 20, value = 1) # Slider for eta
plotOutput("cummPlot2") # Plot output
```

```{r, context="server"}
# Plot the cumulative distribution function
output$cummPlot2 <- renderPlot({
  x <- seq(0, 50, by = .1)
  plot(pweibull(x, shape = 1, scale = input$eta2), col = "red", type = "l", xlab = "Variable", ylab = "Cumulative Probability") 
})
```

As you adjust Eta, notice how the scale of the Weibull changes. Increasing Eta 
shifts the curve to the right, indicating a longer lifespan. Decreasing Eta shifts the
curve to the left, indicating a shorter lifespan. 

## WeibullR

### Getting Started with WeibullR

For this tutorial, we will use WeibullR: an R package for Weibull analysis.

First, check if WeibullR is installed in R and install if not.

```{r echo=TRUE, message=FALSE}
if(!(require(WeibullR))){install.packages('WeibullR')}
```

As an example, suppose a factory has 5 machines that fail at different times. The factory wants to understand the reliability of the machines over time and predict when future failures may occur. The factory collects the failure data and uses Weibull analysis to model the reliability of the machines. The failure times are 30, 49, 82, 90, and 96 respectively. 

To run Weibull analysis in R, first create a vector with the failure data, then use the `MLEw2p` function to fit a Weibull to the data.

```{r echo=TRUE}
failures<-c(30, 49, 82, 90, 96)
fit<-MLEw2p(failures, bounds=TRUE, show=TRUE)
```

### The Weibull Probability Plot

The Weibull Probability Plot shows the time (or equivalent) on the horizontal axis and the unreliability on the vertical axis. In this case, unreliability is analogous to the cumulative probability of failure. We know from the cumulative distribution function that the opposite of the failure probability is the survival probability or reliability. For example, if the unreliability at time 50 is 20%, then the reliability at time 50 is 80%.

Note that the horizontal axis on a probability plot is log scale, and the vertical axis is log(log(1-p)) scale, where p=(i-0.3)/(n+4) and i is the rank of the observation. By using this scale, the resulting plot of the Weibull data appears linear.

```{r quiz2, echo=FALSE}
quiz(caption = "Quiz: Interpreting the Probability Plot",
  question("What is the approximate unreliability at time 100?",
    answer("20%", message = "20% is too low."),
    answer("80%", message = "80% is slightly too low."),
    answer("10%", message = "10% is the reliability, not the unreliability."),
    answer("90%", correct = TRUE, message = "Correct! The unreliability at time 100 is approximately 90%."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("What is the approximate reliability at time 50?)",
    answer("20%", correct = TRUE, message = "Correct! The reliability at time 50 is approximately 20%."),
    answer("80%", message = "80% is the unreliability, not the reliability."),
    answer("10%", message = "10% is slightly too low."),
    answer("90%", message = "90% is too high."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("What is the approximate beta of the fit?",
    answer("1.5", message = "1.5 is too low."),
    answer("3.5", message = "3.5 is slightly too high."),
    answer("0.5", message = "0.5 is too low."),
    answer("3.2", correct = TRUE, message = "Correct! The beta of the fit is approximately 1.5."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("What is the approximate eta of the fit?",
    answer("77.81", correct = TRUE, message = "Correct! The eta of the fit is approximately 77.81."),
    answer("88.84", message = "88.84 is slightly too high."),
    answer("37.89", message = "37.89 is too low."),
    answer("389.7", message = "389.7 is too high."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Data Censoring

In life data analysis, data is often **censored**. Censoring occurs when the exact time to failure is not known for all units.

### Right Censored Data

Right censored data includes **suspensions** or units that have operated for a period of time without failure. 

Now suppose that the factory has 3 additional machines that did not fail during the analysis period. The machines were suspended (i.e. removed from service) at times 100, 45, and 10 respectively.

To add suspensions to the previous example, create a vector with the suspension data and add it to the `MLEw2p` function.

```{r echo=TRUE}
suspensions<-c(100, 45, 10)
fit<-MLEw2p(failures, suspensions, bounds=TRUE, show=TRUE)
```

Note how the fit has changed after adding suspensions. The fit now has a lower beta and higher eta. 

### Interval Censored Data

Interval censored data is most commonly in the form of inspection data, where the failure occurred between two inspection times, the latest known working time (left entry), and the inspection time when the failure was detected (right entry). In this case, the exact failure times are unknown, but are known to have occurred within the intervals. 

To illustrate interval censored data, let's look at an example from Silkworth, 2020. The data consists of part cracks detected during periodic inspections. The left and right columns represent the inspection intervals, and the qty column represents the number of cracks detected in each interval. 

```{r echo=TRUE}
inspection_data<-data.frame(left=c(0, 6.12, 19.92, 29.64, 35.4, 39.72, 45.32, 52.32),
                            right=c(6.12, 19.92, 29.64, 35.4, 39.72, 45.32, 52.32, 63.48),
                            qty=c(5, 16, 12, 18, 18, 2, 6, 17))
```

A number of units did not fail during the inspection period. These units are considered suspensions and are added to the data set as a single entry with the time equal to the last inspection time, event equal to 0 (suspension), and qty equal to the number of suspensions.

```{r echo=TRUE}
suspensions<-data.frame(time=63.48, event=0, qty=73)
```

To run the interval censored example in WeibullR, use the `wblr` function with the `interval` argument to specify the inspection data. Then use the `wblr.fit` and `wblr.conf` functions to fit the model and add confidence bounds. Finally, use the `plot` function to create the probability plot.

```{r echo=TRUE, warning=FALSE}
obj1<-wblr(suspensions, interval = inspection_data)
obj1<-wblr.fit(obj1, method.fit="mle", col="red") 
obj1<-wblr.conf(obj1, method.conf="fm", lty=2)
plot(obj1)
```

Instead of points, the interval censored data is shown as black horizontal lines representing the intervals where failures occurred.

### Grouped Data

Life data is often compiled into groups (i.e. a group of machines failed or survived at a certain time). 

Notice how the previous example contained groups in the `qty` column. Let's look at the inspection data again. 

``` {r echo=TRUE}
knitr::kable(
  head(inspection_data, topn=10)
)
```

The `qty` column indicates how many units failed in each interval. For example, 5 units failed between time 0 and 6.12, 16 units failed between time 6.12 and 19.92, and so on. 

```{r quiz3, echo=FALSE}
quiz(caption = "Quiz: Interval Censored Data",
  question("In the interval censored example above, how many units are in the group of suspensions that did not fail?",
    answer("73", correct = TRUE, message = "Correct! There are 73 units in the group of suspensions that did not fail."),
    answer("17", message = "17 is the number of failures in the last interval."),
    answer("94", message = "94 is the total number of units (failures + suspensions)."),
    answer("63.48", message = "63.48 is the time of the last inspection."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("In the interval censored example, how many total failures occurred?",
    answer("73", message = "73 is the number of suspensions that did not fail."),
    answer("17", message = "17 is the number of failures in the last interval."),
    answer("94", correct = TRUE, message = "Correct! There are a total of 94 failures."),
    answer("63.48", message = "63.48 is the time of the last inspection."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("In the interval censored example, how many total units are in the data set?",
    answer("73", message = "73 is the number of suspensions that did not fail."),
    answer("17", message = "17 is the number of failures in the last interval."),
    answer("94", message = "94 is the total number of failures."),
    answer("167", correct = TRUE, message = "Correct! There are a total of 167 units (failures + suspensions)."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("In the interval censored example, how many failures occurred between time 19.92 and 35.4?",
    answer("12", message = "12 is the number of failures between time 6.12 and 19.92."),
    answer("18", correct = TRUE, message = "Correct! There are 18 failures between time 19.92 and 35.4."),
    answer("16", message = "16 is the number of failures between time 0 and 6.12."),
    answer("2", message = "2 is the number of failures between time 39.72 and 45.32."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Parameter Estimation Methods

### Maximum Likelihood vs Rank Regression

-   **Median Rank Regression (MRR)** is a least squares analysis based on minimizing the squared distance from the cumulative probability to the fitted model.

-   **Maximum Likelihood Estimation (MLE)** is based on estimating the parameters of the distribution by maximizing the likelihood function of the observed data.

WeibullR includes functions for both methods, such as the `MRRw2p`, `MRRw3p` functions for MRR and the `MLEw2p`, `MLEw3p` functions for MLE.

To illustrate the differences between MRR and MLE, let's rerun the right censored example and compare the MRR and MLE methods using the `wblr` function.

```{r echo=TRUE}
MRRfit <- wblr.fit(wblr(failures, suspensions, col="blue"), method.fit="rr")
MLEfit <- wblr.fit(wblr(failures, suspensions, col="red"), method.fit="mle")
plot.wblr(list(MRRfit, MLEfit))
```

The blue model is the MRR fit and the red model is the MLE fit. While both models fit the data well,
there are some slight differences in the parameters and unreliability.

```{r quiz4, echo=FALSE}
quiz(caption = "Quiz: MRR vs MLE",
  question("What is the difference between the MRR and MLE fits (select all that apply)?",
    answer("The MLE fit has a higher beta", correct = TRUE),
    answer("The MRR fit has a higher beta"),
    answer("The MLE fit results in a higher unreliability at time 30"),
    answer("The MRR fit results in a higher unreliability at time 30", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("Which fit has a slightly higher unreliability at time 100?",
    answer("MRR", message = "The MRR fit has a slightly lower unreliability at time 100."),
    answer("MLE", correct = TRUE, message = "The MLE fit has a slightly higher unreliability at time 100."),
    answer("Both are the same", message = "The fits are slightly different at time 100."),
    answer("Cannot be determined from the plot", message = "The unreliability can be determined from the plot."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Other Weibull Models

### The WeiBayes Model

A WeiBayes or 1P Weibull has a fixed $\beta$ or shape parameter based on *prior* knowledge or experience. In this case, $\beta$ is often taken from another Weibull model. 

Weibayes models are typically used when failure data is limited and/or other Weibull models do not fit the data well. 

As an example, let's rerun the right censored example of a machine that failed at times 30, 49, 82, 90, and 96. In this case, the data is limited, so a WeiBayes model may be appropriate. First, recreate the failure and suspension data. Then fit the WeiBayes model while assuming a Beta of 2 and plot the results.

```{r echo=TRUE, warning=FALSE}
failures<-c(30, 49, 82, 90, 96)
obj <- wblr.fit(wblr(failures, suspensions, col="blue"), method.fit="weibayes", weibayes.beta=2)
plot(obj)
```

Note that confidence bounds are not shown for the WeiBayes model. Confidence bounds for the WeiBayes model are not well defined because the shape parameter is fixed.

### The 3-Parameter Weibull Model

The 3P Weibull has an additional parameter *t~0~* to represent a failure free period (i.e. a machine cannot fail until after time *t~0~*). Often parts or machines must age, wear, fatigue, etc. before they can actually fail, and the 3P Weibull can model this behavior.

$$ R(t) = 1 - F(t) = e^-((t-t~0~)/\eta)^\beta $$
Here is an example from "The New Weibull Handbook" by Robert Abernethy (2004). This example consists of 25 failure times from a machine that has a failure free period. Let's run this example as a 3P Weibull. First, create the failure data. Then fit the 3P Weibull model and plot the results.

```{r echo=TRUE}
failures<-c(3.46623, 3.732711, 4.052996, 4.628703, 4.8157, 5.84517, 5.888313, 5.892967,
            8.168362, 10.02799, 10.06062, 10.49785, 11.11493, 11.87369, 12.21122, 12.51854, 
            12.91357, 18.04246, 18.20712, 19.57305, 21.20873, 30.03917, 34.88001, 36.87355,
            53.91168)
fit<- wblr.conf(wblr.fit(wblr(failures), dist="weibull3p"), col="darkgreen")
plot(fit)
```

Note that the 3P Weibull model has a failure free period of approximately 3.3 time units.
The model appears curved at lower probabilities, which is typical of a 3P Weibull.
The curvature is due to the failure free period, which shifts the distribution to the right.

```{r quiz5, echo=FALSE}
quiz(caption = "Quiz: Other Weibull Models",
  question("In the Weibayes example, what is eta?",
    answer("88.84", correct = TRUE, message = "Correct! Eta is approximately 88.84."),
    answer("389.7", message = "389.7 is too high."),
    answer("87.89", message = "87.89 is slightly too low."),
    answer("37.89", message = "37.89 is too low."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("What is the failure free period in the 3P Weibull example?",
    answer("3.3", correct = TRUE, message = "Correct! The failure free period is approximately 3.3 time units."),
    answer("0.85", message = "0.85 is the beta parameter, not the failure free period."),
    answer("11.16", message = "11.16 is the eta parameter, not the failure free period."),
    answer("0", message = "The failure free period is greater than 0."),
    random_answer_order = TRUE
  ),
  question("In the 3P Weibull example, what is beta?",
    answer("0.85", correct = TRUE, message = "Correct! Beta is approximately 0.85."),
    answer("3.3", message = "3.3 is the failure free period, not beta."),
    answer("11.16", message = "11.16 is the eta parameter, not beta."),
    answer("1", message = "Beta is not equal to 1."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("In the 3P Weibull example, what is eta?",
    answer("11.16", correct = TRUE, message = "Correct! Eta is approximately 11.16."),
    answer("0.85", message = "0.85 is the beta parameter, not eta."),
    answer("3.3", message = "3.3 is the failure free period, not eta."),
    answer("1", message = "Eta is not equal to 1."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Multi-Plots

### Multi-Plots

Multi-plots are useful for comparing multiple Weibull models on the same plot. To build a multi-plot in WeibullR, create multiple `wblr` objects and add them to a list. Then use the `plot.wblr` function to create the multi-plot.

First, recreate the failure data from the first example of a machine that failed at times 30, 49, 82, 90, and 96.

```{r echo=TRUE}
failures<-c(30, 49, 82, 90, 96)
obj1<-wblr.fit(wblr(failures), col="red")
```

Next, recreate the failure and suspension data from the right censored example of a machine that failed at times 30, 49, 82, 90, and 96 with suspensions at times 100, 45, and 10.

```{r echo=TRUE}
failures<-c(30, 49, 82, 90, 96)
suspensions<-c(100, 45, 10)
obj2<-wblr.fit(wblr(failures, suspensions, col="purple"))
```

Finally, add the 2 `wblr` objects to a list and plot both objects in a single chart.

```{r echo=TRUE}
plot.wblr(list(obj1, obj2))
```

The red model is the fit without suspensions and the purple model is the fit with suspensions. 
Note how the fit with suspensions has a lower beta and higher eta.

```{r quiz6, echo=FALSE}
quiz(caption = "Quiz: Multi-Plots",
  question("What is the difference between obj1 and obj2 (select all that apply)?",
    answer("obj1 includes suspensions"),
    answer("obj1 has a fit with a lower beta"),
    answer("obj2 includes suspensions", correct=TRUE),
    answer("obj2 has a fit with a lower beta", correct=TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("Which fit has a higher unreliability at time 100?",
    answer("obj2", message = "obj2 has a slightly lower unreliability at time 100."),
    answer("obj1", correct = TRUE, message = "Correct! obj1 has a slightly higher unreliability at time 100."),
    answer("Both are the same", message = "The fits are slightly different at time 100."),
    answer("Cannot be determined from the plot", message = "The unreliability can be determined from the plot."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Competing Failure Modes

A part or a system may fail in different ways. A failure mode is a specific way in which a part or machine can fail. For example, a machine may fail due to corrosion, fatigue, or overload. Each failure mode may have a different distribution of time to failure. Competing failure modes occur when multiple failure modes are present in the same data set. 

When competing failure modes are present, a basic Weibull model tends to fit the data poorly. This is because the basic Weibull assumes that all failures are due to a single failure mode. 

To illustrate, let's simulate some data with 3 different failure modes, where each failure mode has a different beta and eta. 

```{r echo=TRUE}
set.seed(123)
data <- data.frame(
  time = c(rweibull(5, 0.5, 20), rweibull(10, 1, 10), rweibull(5, 2, 5), rweibull(100, 2, 10)),
  event = c(rep(1, 20), rep(0, 100)),
  failure_mode = c(rep("A", 5), rep("B", 10), rep("C", 5), rep("", 100))
  )
```

Next, fit a basic Weibull model to the combined data and plot the results.

```{r echo=TRUE}
obj <- wblr.conf(wblr.fit(wblr(data)))
plot(obj, col="darkgreen", is.plot.legend = FALSE)
```

When the data is combined, the overall fit has a beta of approximately 1, which indicates random failures. However, we know that the data actually contains 3 different failure modes.

Now let's create a separate data set for each failure mode.

```{r echo=TRUE}
dat1 <- data
dat2 <- data
dat3 <- data
dat1$event[dat1$failure_mode != "A"] <- 0
dat2$event[dat2$failure_mode != "B"] <- 0
dat3$event[dat3$failure_mode != "C"] <- 0
```

Finally, fit a `wblr` object for each failure mode and create a multi-plot.

```{r echo=TRUE}
obj1 <- wblr.fit(wblr(dat1, col="blue"))
obj2 <- wblr.fit(wblr(dat2, col="red"))
obj3 <- wblr.fit(wblr(dat3, col="orange"))
plot.wblr(list(obj1, obj2, obj3), is.plot.legend = FALSE)
```

Note how each failure mode has a different beta and eta. Failure mode A (blue) has a beta less than 1, which means the failure rate is decreasing. Failure mode B (red) has a beta greater than 1, which means that the failure rate is increasing. Failure mode C (orange) has a beta close to 1, which means that the failure rate is approximately constant.

## Contour Plots

Another way to compare Weibull models is with contour plots. A contour plot is a graphical representation of the relationship between the Weibull parameters Beta and Eta. 
 
Once again, let's look at the first example of a machine that failed at times 30, 49, 82, 90, and 96. First, recreate the failure data. Then fit the Weibull model and build the contour plot at a probability level of 90%.

```{r echo=TRUE, results='hide'}
failures<-c(30, 49, 82, 90, 96)
obj <- wblr.conf(wblr.fit(wblr(failures, col="blue"), method="mle"), method="lrb")
plot_contour(obj, CL=0.9)
```

The Weibull contour plot shows the confidence region for the Weibull parameters Beta and Eta at a probability level of 90%. The x-axis represents Eta, the y-axis represents Beta, and the contour lines represent the confidence levels. 

### Comparing Contour Plots

To compare multiple Weibull models, we can build contour plots for each model and overlay them on the same chart. If the models are similar, then the contours will overlap. If the models are different, then the contours will not overlap.

To visualize this concept, let's revisit the previous competing failure mode example. First, recreate the failure data with 3 different failure modes. Then create a separate `wblr` object for each failure mode and build the contour plot for each model at a probability level of 90%.


```{r echo=TRUE, results='hide'}
obj1 <- wblr.conf(wblr.fit(wblr(dat1, col="blue"), method="mle"), method="lrb")
obj2 <- wblr.conf(wblr.fit(wblr(dat2, col="red"), method="mle"), method="lrb")
obj3 <- wblr.conf(wblr.fit(wblr(dat3, col="orange"), method="mle"), method="lrb")
plot_contour(list(obj1, obj2, obj3), CL=0.9)
```

The blue, red, and orange contours represent the 3 different failure modes. Note how the contours overlap, which indicates that the failure modes are not significantly different. If the contours did not overlap, then the failure modes would be significantly different.

```{r quiz7, echo=FALSE}
quiz(caption = "Quiz: Contour Plots",
  question("In the single contour plot example, what is the approximate beta at the center of the contour?",
    answer("Between 2 and 3", message = "3 is slightly too low."),
    answer("Between 75 and 80", message = "The approximate eta is between 75 and 80."),
    answer("Between 4 and 4.5", message = "4 is slightly too high."),
    answer("Between 3 and 3.5", correct = TRUE, message = "Correct! The approximate beta at the center of the contour is between 3 and 3.5."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("In the multi-contour plot example, at a probability level of 90%...",
    answer("obj3 (orange) is significantly different than the others", message = "obj3 (orange) overlaps with obj1 (blue) and obj2 (red), so they are not significantly different."),
    answer("obj2 (red) is significantly different than the others", message = "obj2 (red) overlaps with obj3 (orange), so they are not significantly different."),
    answer("None of the distributions are significantly different", correct=TRUE, message = "Correct! All of the distributions overlap, so they are not significantly different."),
    answer("All of the distributions are significantly different", message = "The distributions overlap, so they are not significantly different."),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Summary

Congratulations on completing the tutorial on life data analysis!

In this tutorial, we've introduced the concepts of life data analysis and reliability. You have learned to calculate reliability using the Weibull distribution and visualize life data.

### To Get Help

To get help with a specific function in WeibullR, type a question mark before the function name.

```{r echo=TRUE, eval=FALSE}
?MRRw2p
```

For more help with the WeibullR package.

```{r echo=TRUE, eval=FALSE}
help(package="WeibullR")
```

### References

- Abernethy, R.B. (2004) The New Weibull Handbook. Fifth Edition.

-   Aden-Buie G, Schloerke B, Allaire J, Rossell Hayes A (2023). learnr: Interactive Tutorials for R. https://rstudio.github.io/learnr/, https://github.com/rstudio/learnr.

-   Silkworth, David. (2020). WeibullR: An R Package for Weibull Analysis for Reliability Engineers. 43-53. <https://doi.org/10.35566/isdsa2019c3>.

-   Silkworth D, Symynck J (2022). *WeibullR: Weibull Analysis for Reliability Engineering*. R package version 1.2.1, <https://CRAN.R-project.org/package=WeibullR>.
