# Reliability Growth Analysis {#sec-rt}

## Introduction

**Reliability Growth Analysis (RGA)** monitors and improves reliability over time by analyzing failure data collected during testing. It identifies trends in the failure rate, confirming whether corrective actions are having the intended effect.

## Learning Objectives

By the end of this chapter, you will be able to:

- Define key reliability growth concepts: Crow-AMSAA and Duane models.
- Use `rdt()` to plan a reliability demonstration test (required test time and sample size).
- Fit reliability growth models using R.
- Apply the Crow-AMSAA model to assess reliability growth.
- Use piecewise NHPP models to detect design changes during testing.
- Forecast future failures from a fitted growth model using `predict_rga()`.
- Interpret reliability growth plots and make decisions from the results.

## Reliability Growth Analysis

The `ReliaGrowR` package [@ReliaGrowR] provides functions for reliability growth analysis in R.

```{r}
#| message: false
library(ReliaGrowR)
```

## Test Planning

Before analyzing test data, it helps to answer two design questions: *How long must we test?* and *How many failures should we expect?* ReliaGrowR provides tools for both.

### Reliability Demonstration Tests — `rdt()`

A **Reliability Demonstration Test (RDT)** is a pass/fail test designed to demonstrate that a product meets a reliability target at a specified confidence level. `rdt()` calculates the required test time given the sample size (or vice versa).

```{r}
# How long must we test 10 units to demonstrate R(1000 hrs) >= 0.90
# at 80% confidence? Assume Weibull shape beta = 2, zero allowable failures.
plan <- rdt(
  target       = 0.90,   # target reliability
  mission_time = 1000,   # mission time (hours)
  conf_level   = 0.80,   # confidence level
  beta         = 2,      # assumed Weibull shape parameter
  n            = 10,     # number of test units
  f            = 0       # allowable failures
)
print(plan)
```

Each unit must be tested for approximately 1,236 hours without failure. If the test passes, we have demonstrated $R(1000) \geq 0.90$ with 80% confidence.

::: {.callout-note}
## Try It

Compare the required test time for sample sizes of 5, 10, and 20 units (keep all other parameters the same). What is the trade-off?

```{r}
# Vary n to see the effect on required test time
for (n in c(5, 10, 20)) {
  # plan <- rdt(target = 0.90, mission_time = 1000, conf_level = 0.80,
  #             beta = 2, n = n, f = 0)
  # print(plan)
}
```

<details><summary>Solution</summary>
```{r}
for (n in c(5, 10, 20)) {
  plan <- rdt(target = 0.90, mission_time = 1000, conf_level = 0.80,
              beta = 2, n = n, f = 0)
  print(plan)
}
```

More test units means shorter required test time per unit — but total test effort (n × T) stays roughly constant. Larger fleets reduce calendar time at the cost of more physical units.
</details>
:::

### Forecasting Growth — `predict_rga()`

Once a growth test is underway, `predict_rga()` projects cumulative failures beyond the current endpoint — useful for deciding whether to continue testing or whether a reliability target will be met by a program deadline.

```{r}
times    <- c(100, 200, 300, 400, 500)
failures <- c(1, 2, 1, 3, 2)

fit  <- rga(times, failures)
pred <- predict_rga(fit, times = seq(500, 1000, by = 100))
print(pred)
```

The lower and upper bounds reflect the uncertainty in the Crow-AMSAA parameter estimates. A narrow interval indicates the growth trend is well characterized; a wide interval suggests more test time is needed.

::: {.callout-tip}
## Review

The fitted model (using the Crow-AMSAA framework, described fully below) gives $\hat{\beta} = 0.72$. Is the system still improving at time 1,000?

<details><summary>Answer</summary>
Yes — for the Crow-AMSAA model, $\beta < 1$ means the cumulative failure rate is **decreasing** (the growth curve is concave), so reliability is still improving at time 1,000. Testing should continue until $\beta$ stabilizes near the growth target.
</details>
:::

## The Duane Model

The **Duane Model** [@Duane1964] is one of the earliest and most widely used models for reliability growth. It is a log-log plot of cumulative MTBF vs cumulative time, where the MTBF is the total operating time divided by the number of failures up to that time.

$$\text{CMTBF}(t) = K \cdot t^{\beta - 1}$$

where $K > 0$ is a scale parameter estimated from the data and $\beta$ is the growth slope. The slope of the line on the log-log plot indicates the rate of reliability growth:

| Slope ($\beta$) | Meaning |
|:---:|---|
| $> 1$ | Reliability improving (failure rate decreasing) |
| $= 1$ | No change (stable) |
| $< 1$ | Reliability worsening (failure rate increasing) |

Plotting the Duane model for three beta values:

```{r}
t <- seq(1, 100, by = 0.1)

# Three scenarios
plot(log(t), log(0.1 * t^(1.5 - 1)), type = "l", col = "blue", lwd = 2,
     xlab = "log(Cumulative Time)", ylab = "log(Cumulative MTBF)",
     main = "Duane Plot: Three Reliability Growth Scenarios",
     ylim = c(-4, 1))
lines(log(t), log(0.1 * t^(1.0 - 1)), col = "red",       lwd = 2)
lines(log(t), log(0.1 * t^(0.6 - 1)), col = "darkgreen", lwd = 2)
legend("topleft",
       legend = c("β = 1.5 (improving)", "β = 1.0 (stable)", "β = 0.6 (worsening)"),
       col = c("blue", "red", "darkgreen"), lwd = 2)
```

### Fitting with ReliaGrowR

```{r}
times    <- c(100, 200, 300, 400, 500)
failures <- c(1, 2, 1, 3, 2)

fit <- duane(times, failures)
plot(fit, main = "Duane Model Example",
     xlab = "Cumulative Time", ylab = "Cumulative MTBF")
```

::: {.callout-note}
## Try It

A new system was tested and the following cumulative failure counts were recorded. Fit a Duane model and plot the result.

```{r}
times    <- c(200, 450, 750, 1100, 1500, 2000)
failures <- c(3, 2, 2, 1, 1, 1)
# fit <- duane(times, failures)
# plot(fit)
```

<details><summary>Solution</summary>
```{r}
times    <- c(200, 450, 750, 1100, 1500, 2000)
failures <- c(3, 2, 2, 1, 1, 1)
fit <- duane(times, failures)
plot(fit, main = "Duane Reliability Growth Plot",
     xlab = "Cumulative Test Time", ylab = "Cumulative MTBF")
```
</details>
:::

::: {.callout-tip}
## Review

What does a Duane plot with a slope greater than 1 indicate?

<details><summary>Answer</summary>
The system's reliability is **improving** — the failure rate is decreasing over time.
</details>
:::

## The Crow-AMSAA Model

The **Crow-AMSAA Model** [@Crow1975] models cumulative failures vs cumulative time using a Non-Homogeneous Poisson Process (NHPP):

$$N(t) = \lambda_0 \cdot t^{\beta}$$

where $\lambda_0 > 0$ is the scale parameter and $\beta$ is the shape parameter.

The shape parameter $\beta$ interpretation:

| $\beta$ | Meaning |
|:---:|---|
| $> 1$ | Failures increasing (reliability worsening) |
| $= 1$ | Constant rate (stable) |
| $< 1$ | Failures decreasing (reliability improving) |

::: {.callout-warning}
## β Interpretation is Reversed Between Models

The Crow-AMSAA $\beta$ means the **opposite** of the Duane $\beta$:

- **Duane**: $\beta > 1$ = reliability *improving* (MTBF is rising).
- **Crow-AMSAA**: $\beta > 1$ = reliability *worsening* (cumulative failures are accelerating).

This reversal occurs because Duane plots MTBF (higher is better), while Crow-AMSAA plots cumulative failures (a lower slope means fewer failures, so *improving*). Always check which model produced your $\beta$ before interpreting the result.
:::

```{r}
t <- seq(0.1, 100, by = 0.1)

# Three beta scenarios for cumulative failures
plot(t, 0.5 * t^0.6, type = "l", col = "blue", lwd = 2,
     xlab = "Cumulative Time", ylab = "Cumulative Failures",
     main = "Crow-AMSAA: Three Reliability Scenarios",
     ylim = c(0, 20))
lines(t, 0.5 * t^1.0, col = "red",       lwd = 2)
lines(t, 0.5 * t^1.5, col = "darkgreen", lwd = 2)
legend("topleft",
       legend = c("β = 0.6 (improving)", "β = 1.0 (stable)", "β = 1.5 (worsening)"),
       col = c("blue", "red", "darkgreen"), lwd = 2)
```

### Fitting with ReliaGrowR

```{r}
result <- rga(times, failures)
plot(result, main = "Crow-AMSAA Model Example",
     xlab = "Cumulative Time", ylab = "Cumulative Failures")
```

::: {.callout-note}
## Try It

A development test recorded these cumulative failure counts. Use `rga()` with `model_type = "Crow-AMSAA"` to fit the model.

```{r}
times    <- c(50, 120, 220, 360, 530, 740, 1000)
failures <- c(4, 3, 3, 2, 2, 1, 1)
# fit <- rga(times, failures, model_type = "Crow-AMSAA")
# plot(fit)
```

<details><summary>Solution</summary>
```{r}
times    <- c(50, 120, 220, 360, 530, 740, 1000)
failures <- c(4, 3, 3, 2, 2, 1, 1)
fit <- rga(times, failures, model_type = "Crow-AMSAA")
plot(fit, main = "Crow-AMSAA Reliability Growth",
     xlab = "Cumulative Test Time", ylab = "Cumulative Failures")
```
</details>
:::

### The Piecewise NHPP Model

The **Piecewise NHPP** fits a separate Power Law to each segment of the time axis, separated by breakpoints. This is useful when a design change occurs during testing.

$$N_i(t) = N(t_{i-1}) + \lambda_i (t - t_{i-1})^{\beta_i}, \quad t_{i-1} < t \leq t_i$$

```{r}
times    <- c(25, 55, 97, 146, 201, 268, 341, 423, 513, 609, 710, 820, 940, 1072, 1217)
failures <- c(1, 1, 2, 4, 4, 1, 1, 2, 1, 4, 1, 1, 3, 3, 4)
breaks   <- 500

result <- rga(times, failures, model_type = "Piecewise NHPP", breaks = breaks)
plot(result, main = "Piecewise NHPP Model",
     xlab = "Cumulative Time", ylab = "Cumulative Failures")
```

The two segments have different slopes — the first segment (pre-change) has a higher slope than the second (post-change), confirming the design change reduced the failure rate.

### Change Point Detection

Without a known breakpoint, `rga()` can detect the change point automatically:

```{r}
result_auto <- rga(times, failures, model_type = "Piecewise NHPP")
plot(result_auto, main = "Piecewise NHPP with Change Point Detection",
     xlab = "Cumulative Time", ylab = "Cumulative Failures")
```

The automatically detected change point is near time 500, consistent with the known design change.

## Summary

**Key takeaways:**

- **Duane model**: log-log of MTBF vs time; positive slope ($\beta > 1$) = improving.
- **Crow-AMSAA**: log of cumulative failures vs time; concave downward = improving ($\beta < 1$).
- **Piecewise NHPP**: fits multiple segments; use when a known change occurs during testing.
- For accelerated life testing (Arrhenius, Power Law, relationship plots), see @sec-alt.

