# Repairable Systems Analysis {#sec-rs}

## Introduction

Unlike life data analysis (@sec-lda), which focuses on the *time to first failure* of non-repairable components, **Repairable Systems Analysis** [@Ascher1984; @Meeker1998] models the entire sequence of recurring failures on a single system or fleet. This chapter covers the foundational theory and the `ReliaGrowR` tools for fitting models, computing fleet-level exposure, and estimating the Mean Cumulative Function.

## Learning Objectives

By the end of this chapter, you will be able to:

- Define the distinction between repairable and non-repairable systems.
- Explain the Homogeneous Poisson Process (HPP) and Non-Homogeneous Poisson Process (NHPP).
- Describe the Power Law Process and its parameters.
- Fit and interpret the Crow-AMSAA model using `nhpp()` from ReliaGrowR.
- Apply the Piecewise NHPP model to detect changes in failure behavior.
- Compute fleet-level exposure using `exposure()`.
- Estimate the Mean Cumulative Function (MCF) using `mcf()`.
- Forecast future failure counts from a fitted Power Law model.

The `ReliaGrowR` package [@ReliaGrowR] provides functions for repairable systems analysis in R.

```{r}
#| message: false
library(ReliaGrowR)
```

## Repairable vs Non-Repairable Systems

A **non-repairable system** is discarded after failure and replaced with a new, statistically identical unit. The unit of analysis is the *time to first failure*. Weibull analysis applies.

A **repairable system** is restored to service after each failure and continues accumulating operating time. The unit of analysis is the *sequence of inter-failure times* or cumulative failure counts over the system's life.

Two repair assumptions are common:

| Repair type | Description | Model |
|---|---|---|
| Perfect repair | As-good-as-new after each repair | Renewal process |
| Minimal repair | As-bad-as-old — failure fixed, nothing else changed | NHPP |

Most field maintenance data best fits the **minimal repair** assumption, making NHPP the standard framework.

The **recurrence rate** (intensity function) $\rho(t)$ is the instantaneous rate of failure occurrence at time $t$:

$$\rho(t) = \frac{d\,E[N(t)]}{dt}$$

## Failure Processes: HPP and NHPP

### Homogeneous Poisson Process (HPP)

The **HPP** assumes a *constant* failure rate $\lambda$ throughout the system's life:

$$E[N(t)] = \lambda t \qquad \rho(t) = \lambda \qquad \text{MTBF} = 1/\lambda$$

The HPP is appropriate when failures occur at random with no trend.

### Non-Homogeneous Poisson Process (NHPP)

The **NHPP** allows the intensity function to change with time:

$$\rho(t) = \frac{d\Lambda(t)}{dt}$$

When $\rho(t)$ increases, the system is deteriorating; when it decreases, the system is improving.

```{r}
t <- seq(1, 100, by = 1)

# HPP (constant rate)
hpp_intensity  <- rep(0.05, length(t))

# NHPP Power Law scenarios
nhpp_improving  <- 0.01 * 0.7 * t^(0.7 - 1)   # beta = 0.7
nhpp_stable     <- 0.05 * 1.0 * t^(1.0 - 1)   # beta = 1.0
nhpp_worsening  <- 0.002 * 1.5 * t^(1.5 - 1)  # beta = 1.5

ylim_max <- max(c(hpp_intensity, nhpp_improving, nhpp_stable, nhpp_worsening)) * 1.1

plot(t, nhpp_improving, type = "l", col = "blue", lwd = 2,
     ylim = c(0, ylim_max),
     xlab = "Time", ylab = "Intensity ρ(t)",
     main = "NHPP Power Law Intensity Functions")
lines(t, nhpp_stable,    col = "red",       lwd = 2)
lines(t, nhpp_worsening, col = "darkgreen", lwd = 2)
abline(h = 0.05, col = "gray50", lwd = 2, lty = 2)
legend("topright",
       legend = c("β = 0.7 (improving)", "β = 1.0 (stable / HPP)",
                  "β = 1.5 (worsening)", "HPP reference λ = 0.05"),
       col = c("blue", "red", "darkgreen", "gray50"),
       lwd = 2, lty = c(1, 1, 1, 2))
```

## The Power Law Process

The **Power Law Process** is the most widely used NHPP for repairable systems:

$$E[N(t)] = \lambda t^{\beta} \qquad \rho(t) = \lambda \beta t^{\beta-1}$$

**Parameter interpretation:**

- $\lambda > 0$ — scale parameter; overall magnitude of the failure rate.
- $\beta < 1$: decreasing intensity (system improving).
- $\beta = 1$: constant intensity = HPP.
- $\beta > 1$: increasing intensity (system deteriorating).

The **instantaneous MTBF** at time $t$:

$$\text{MTBF}(t) = \frac{1}{\rho(t)} = \frac{1}{\lambda\beta\,t^{\beta-1}}$$

Maximum likelihood estimates given $n$ failures at times $t_1 < \cdots < t_n$ over total time $T$:

$$\hat{\beta} = \frac{n}{\displaystyle\sum_{i=1}^{n} \ln(T/t_i)} \qquad \hat{\lambda} = \frac{n}{T^{\hat{\beta}}}$$

::: {.callout-tip}
## Review

A repairable system is observed to have an increasing failure rate over time. Which model is most appropriate?

<details><summary>Answer</summary>
**NHPP Power Law with $\beta > 1$**. An increasing intensity function is characteristic of a deteriorating system. HPP (constant rate) cannot capture this trend.
</details>
:::

## NHPP Analysis with nhpp()

The `nhpp()` function fits the Power Law Process (Crow-AMSAA model) [@Crow1975] to cumulative failure data by maximum likelihood, returning estimates of $\lambda$ and $\beta$.

A curve that bends downward (concave) indicates $\hat{\beta} < 1$ — improving system. Upward bending (convex) indicates $\hat{\beta} > 1$ — deteriorating system.

```{r}
times    <- c(500, 1200, 2100, 3300, 4800, 6500, 8400, 10500, 12800, 15000)
failures <- c(3, 4, 4, 5, 4, 3, 3, 3, 2, 2)

fit <- nhpp(time = times, event = failures)
plot(fit,
     main = "Crow-AMSAA Model: Industrial Pump Fleet",
     xlab = "Cumulative Operating Hours",
     ylab = "Cumulative Failures")
```

The concave curve and $\hat{\beta} < 1$ indicate the pump fleet's failure rate is decreasing — a favorable operating regime.

::: {.callout-note}
## Try It

A compressor fleet recorded these cumulative operating hours and failure counts. Fit a Power Law model using `nhpp()` with `method = "LS"`.

```{r}
times    <- c(1000, 2500, 4000, 5500, 8000)
failures <- c(2, 3, 4, 5, 5)
# fit <- nhpp(time = times, event = failures, method = "LS")
# plot(fit)
```

<details><summary>Solution</summary>
```{r}
times    <- c(1000, 2500, 4000, 5500, 8000)
failures <- c(2, 3, 4, 5, 5)
fit <- nhpp(time = times, event = failures, method = "LS")
plot(fit, main = "Crow-AMSAA: Compressor Fleet",
     xlab = "Cumulative Hours", ylab = "Cumulative Failures")
```
</details>
:::

## Piecewise NHPP

The **Piecewise NHPP** [@Guo2010] fits a separate Power Law to each segment of the time axis, allowing for structural changes in failure behavior (overhauls, design modifications, operational changes):

$$E[N(t)] = \begin{cases} \lambda_1 t^{\beta_1} & 0 < t \leq \tau \\ E[N(\tau)] + \lambda_2(t-\tau)^{\beta_2} & t > \tau \end{cases}$$

Pass a breakpoint estimate via the `breaks` argument. The `segmented` algorithm then optimizes the breakpoint location from the data.

```{r}
times2    <- c(500, 1200, 2100, 3300, 4800, 6500, 8400, 9200, 10100, 11300,
               12800, 14500, 16400, 18500, 20800)
failures2 <- c(3, 4, 4, 5, 6, 6, 5, 3, 3, 2, 2, 2, 2, 1, 1)

fit_pw <- nhpp(time = times2, event = failures2, breaks = 8000, method = "LS")
plot(fit_pw,
     main = "Piecewise NHPP: Fleet Overhaul at Hour 8000",
     xlab = "Cumulative Operating Hours",
     ylab = "Cumulative Failures")
```

The first segment (pre-overhaul, steeper slope) gives way to a flatter or downward-bending second segment, confirming the overhaul reduced the failure rate. The fitted breakpoint may differ slightly from the input value — here it is estimated near hour 9,820.

::: {.callout-tip}
## Review

What is the primary reason to use a Piecewise NHPP instead of a single Crow-AMSAA model?

<details><summary>Answer</summary>
To account for a **known or suspected structural change** in the failure process (overhaul, design change, operational change). A piecewise model fits each segment independently, capturing the change in failure behavior.
</details>
:::

## Exposure Analysis

When analyzing a *fleet*, individual units enter and leave observation at different times. **Exposure** is the total operating time accumulated across all units currently at risk.

The `exposure()` function computes:
- **Cumulative exposure** — total system-time at risk up to each event time.
- **Number at risk** — how many units were under observation at each time.
- **Event rate** — cumulative events divided by cumulative exposure.

```{r}
pump_data <- data.frame(
  id    = c(1, 1, 1,  2, 2, 2,  3, 3,  4, 4, 4, 4,  5, 5),
  time  = c(310, 850, 1620,  420, 1050, 2100,  580, 1890,  240, 710, 1380, 2400,  530, 1740),
  event = c(1, 1, 1,   1, 1, 1,   1, 1,   1, 1, 1, 1,   1, 1)
)

exp_result <- exposure(id = pump_data$id, time = pump_data$time,
                       event = pump_data$event)
plot(exp_result)
```

`exp_result` is a data frame containing three columns: `cum_exposure` (total system-time accumulated across all units at risk), `n_risk` (number of units under observation at each time), and `event_rate` (cumulative events divided by cumulative exposure). The exposure plot shows how total observed system-time accumulates and how the fleet-wide event rate behaves over time.

::: {.callout-tip}
## Review

Why is it important to account for exposure when comparing failure rates across systems with different observation periods?

<details><summary>Answer</summary>
Without normalizing by exposure, a system observed for twice as long will appear to have twice as many failures even if its underlying rate is identical. Exposure provides the denominator — failures per unit time — so rates can be compared fairly.
</details>
:::

## Mean Cumulative Function

The **Mean Cumulative Function (MCF)** is a non-parametric estimate of the expected cumulative number of failures *per system* as a function of time. It makes no distributional assumptions and handles systems with different observation periods.

The **Nelson-Aalen estimator** is a non-parametric method that requires no distributional assumption and naturally handles systems with different observation lengths (administrative censoring):

$$\hat{M}(t) = \sum_{t_j \leq t} \frac{d_j}{n_j}$$

where $d_j$ is the number of failures at time $t_j$ and $n_j$ is the number of systems at risk at $t_j$.

Pass `end_time` to `mcf()` when systems were observed beyond their last recorded failure — otherwise the risk set is deflated too early, inflating the rate estimate.

```{r}
end_times  <- c("1" = 3000, "2" = 3000, "3" = 3000, "4" = 3000, "5" = 3000)

mcf_result <- mcf(id = pump_data$id, time = pump_data$time,
                  event = pump_data$event,
                  end_time = end_times)
plot(mcf_result,
     main = "Mean Cumulative Function: Industrial Pump Fleet",
     xlab = "Operating Hours",
     ylab = "Expected Cumulative Failures per System")
```

The slope of the MCF reveals the recurrence trend: steepening = increasing failure rate; flattening = improving.

::: {.callout-note}
## Try It

Three systems were observed with the following failure times. Each was observed until hour 2000. Compute and plot the MCF.

```{r}
id    <- c(rep("1", 3), rep("2", 4), rep("3", 2))
time  <- c(400, 900, 1500,   300, 700, 1200, 1800,   600, 1400)
event <- rep(1, 9)
end_times2 <- c("1" = 2000, "2" = 2000, "3" = 2000)
# fit <- mcf(id, time, event, end_time = end_times2)
# plot(fit)
```

<details><summary>Solution</summary>
```{r}
id    <- c(rep("1", 3), rep("2", 4), rep("3", 2))
time  <- c(400, 900, 1500,   300, 700, 1200, 1800,   600, 1400)
event <- rep(1, 9)
end_times2 <- c("1" = 2000, "2" = 2000, "3" = 2000)
fit <- mcf(id = id, time = time, event = event, end_time = end_times2)
plot(fit, main = "MCF Example", xlab = "Time (hours)", ylab = "MCF")
```
</details>
:::

## Interpreting Results and Predicting Future Failures

### Reading $\hat{\beta}$

| $\hat{\beta}$ | Interpretation | Typical cause |
|:---:|---|---|
| $< 1$ | Decreasing failure rate | Effective maintenance, infant mortality resolved |
| $= 1$ | Constant failure rate | Random external shocks |
| $> 1$ | Increasing failure rate | Wear-out, accumulating damage, aging |

### Cumulative vs Instantaneous MTBF

- **Cumulative MTBF** = $T / N(T)$ — historical average over all observed failures.
- **Instantaneous MTBF** at time $t$ = $1/\rho(t) = 1/(\hat{\lambda}\hat{\beta}\,t^{\hat{\beta}-1})$ — expected time to the *next* failure.

For $\hat{\beta} > 1$ (aging system), the instantaneous MTBF falls *below* the cumulative average at late times — the system's near-term reliability is worse than its history suggests.

### Predicting Future Failures

Expected additional failures from current time $T$ to a future time $T + \Delta t$:

$$E[\Delta N] = \hat{\lambda}(T + \Delta t)^{\hat{\beta}} - \hat{\lambda}\, T^{\hat{\beta}}$$

```{r}
lambda    <- 0.0003
beta      <- 0.85
T_current <- 15000
T_future  <- 17000

expected_add <- lambda * T_future^beta - lambda * T_current^beta
cat("Expected additional failures:", round(expected_add, 2), "\n")
```

::: {.callout-note}
## Try It

A system has been operating for 5,000 hours. A Power Law fit yielded $\lambda = 0.0005$ and $\beta = 1.2$. Calculate the expected additional failures in the next 1,000 hours.

```{r}
lambda    <- 0.0005
beta      <- 1.2
T_current <- 5000
T_future  <- 6000
# E[ΔN] = lambda * T_future^beta - lambda * T_current^beta
```

<details><summary>Solution</summary>
```{r}
lambda    <- 0.0005
beta      <- 1.2
T_current <- 5000
T_future  <- 6000
expected_failures <- lambda * T_future^beta - lambda * T_current^beta
cat("Expected additional failures:", round(expected_failures, 2), "\n")
```
</details>
:::

::: {.callout-tip}
## Review

A system has a fitted Power Law model with $\beta = 1.35$. What is the operational implication?

<details><summary>Answer</summary>
The failure rate is **increasing** — wear-out or aging is occurring. The system may be approaching a major overhaul threshold. The instantaneous MTBF is below the cumulative historical average.
</details>
:::

## Summary

**Key takeaways:**

- Repairable systems analysis models the *sequence* of recurring failures, not just the first failure.
- The **Power Law Process** ($E[N(t)] = \lambda t^\beta$) is the standard NHPP for repairable systems.
- $\hat{\beta}$ determines whether the failure rate is improving ($< 1$), stable ($= 1$), or worsening ($> 1$).
- `nhpp()` fits the Crow-AMSAA model; add `breaks` for piecewise fitting.
- `exposure()` computes fleet-level operating time at risk.
- `mcf()` estimates cumulative failures per system non-parametrically.
- Future failure counts: $E[\Delta N] = \hat{\lambda}(T + \Delta t)^{\hat{\beta}} - \hat{\lambda}\,T^{\hat{\beta}}$.

