# Interactive Plots with ReliaPlotR {#sec-reliaplotr}

## Introduction

The previous chapters produce static base-R plots. **ReliaPlotR** [@ReliaPlotR] is a companion package that turns those same fitted model objects into *interactive* plotly charts — with hover tooltips, zoom/pan controls, and built-in download buttons — without requiring you to change how you fit your models.

The API follows a single convention: pass the fitted object from `WeibullR`, `ReliaGrowR`, or `WeibullR.ALT` directly to the corresponding `plotly_*` function. The complete function map:

| Function | Input object | Source package | Chapter |
|---|---|---|:---:|
| `plotly_wblr()` | `wblr` | WeibullR | @sec-lda |
| `plotly_contour()` | `wblr` (with `lrb` bounds) | WeibullR | @sec-lda |
| `plotly_duane()` | `duane` | ReliaGrowR | @sec-rt |
| `plotly_rga()` | `rga` | ReliaGrowR | @sec-rt |
| `plotly_alt()` | `alt.parallel` | WeibullR.ALT | @sec-alt |
| `plotly_rel()` | `alt.fit` | WeibullR.ALT | @sec-alt |
| `plotly_nhpp()` | `nhpp` | ReliaGrowR | @sec-rs |
| `plotly_mcf()` | `mcf` | ReliaGrowR | @sec-rs |
| `plotly_exposure()` | `exposure` | ReliaGrowR | @sec-rs |

## Installation

```{r}
#| eval: false
install.packages("ReliaPlotR")
```

```{r}
#| message: false
library(ReliaPlotR)
library(WeibullR)
library(ReliaGrowR)
library(WeibullR.ALT)
```

ReliaPlotR is available on CRAN: <https://CRAN.R-project.org/package=ReliaPlotR>

## Life Data Analysis Plots

These functions accept the `wblr` objects produced by `WeibullR` — exactly as shown in @sec-lda. No reformatting needed.

### Weibull Probability Plot — `plotly_wblr()`

Pass any fitted `wblr` object (or a list of objects for overlays):

```{r}
#| label: fig-wblr-single
#| fig-cap: "Interactive Weibull probability plot with confidence bounds"
failures <- c(30, 49, 82, 90, 96)
obj <- wblr.conf(wblr.fit(wblr(failures)))
plotly_wblr(obj, fitCol = "blue", confCol = "blue",
            main = "Weibull Probability Plot")
```

Hover over any point to read off its exact time and cumulative probability. The results table (β, η, Anderson-Darling) appears in the lower right.

Including suspensions:

```{r}
#| label: fig-wblr-susp
#| fig-cap: "Weibull plot with right-censored suspensions"
suspensions <- c(100, 45, 10)
obj_susp <- wblr.conf(wblr.fit(wblr(failures, suspensions)))
plotly_wblr(obj_susp, susp = suspensions,
            fitCol = "red", confCol = "red",
            main = "Weibull Plot with Suspensions")
```

Overlaying two models — pass a list:

```{r}
#| label: fig-wblr-multi
#| fig-cap: "Multi-model overlay: with and without suspensions"
obj1 <- wblr.conf(wblr.fit(wblr(failures,             col = "red")))
obj2 <- wblr.conf(wblr.fit(wblr(failures, suspensions, col = "purple")))
plotly_wblr(list(obj1, obj2),
            cols  = c("red", "purple"),
            main  = "Multi-Model Weibull Overlay")
```

### Contour Plot — `plotly_contour()`

Pass a `wblr` object fitted with likelihood-ratio bounds (`method.conf = "lrb"`):

```{r}
#| label: fig-contour
#| fig-cap: "Interactive contour plot showing the 90% confidence region for β and η"
obj_lrb <- wblr.conf(wblr.fit(wblr(failures, col = "blue"), method.fit = "mle"),
                     method.conf = "lrb")
plotly_contour(obj_lrb,
               main = "Weibull Contour Plot (90% CL)",
               xlab = "Eta (η)", ylab = "Beta (β)")
```

Hover to read off any (η, β) combination inside the confidence region. Overlaying multiple contours:

```{r}
#| label: fig-contour-multi
#| fig-cap: "Overlaid contour plots for three competing failure modes"
set.seed(123)
data <- data.frame(
  time = c(rweibull(5, 0.5, 20), rweibull(10, 1, 10), rweibull(5, 2, 5),
           rweibull(100, 2, 10)),
  event        = c(rep(1, 20), rep(0, 100)),
  failure_mode = c(rep("A", 5), rep("B", 10), rep("C", 5), rep("", 100))
)
dat1 <- dat2 <- dat3 <- data
dat1$event[dat1$failure_mode != "A"] <- 0
dat2$event[dat2$failure_mode != "B"] <- 0
dat3$event[dat3$failure_mode != "C"] <- 0

c1 <- wblr.conf(wblr.fit(wblr(dat1, col = "blue"),   method.fit = "mle"), method.conf = "lrb")
c2 <- wblr.conf(wblr.fit(wblr(dat2, col = "red"),    method.fit = "mle"), method.conf = "lrb")
c3 <- wblr.conf(wblr.fit(wblr(dat3, col = "orange"), method.fit = "mle"), method.conf = "lrb")

plotly_contour(list(c1, c2, c3),
               cols = c("blue", "red", "orange"),
               main = "Contour Comparison: Three Failure Modes")
```

## Reliability Growth Plots

These functions accept fitted objects from `ReliaGrowR` — the same objects plotted in @sec-rt.

### Duane Plot — `plotly_duane()`

```{r}
#| label: fig-duane
#| fig-cap: "Interactive Duane reliability growth plot"
times    <- c(100, 200, 300, 400, 500)
failures <- c(1, 2, 1, 3, 2)

fit_duane <- duane(times, failures)
plotly_duane(fit_duane,
             main = "Duane Reliability Growth",
             xlab = "Cumulative Time",
             ylab = "Cumulative MTBF")
```

### Crow-AMSAA Plot — `plotly_rga()`

```{r}
#| label: fig-rga
#| fig-cap: "Interactive Crow-AMSAA cumulative failures plot"
fit_rga <- rga(times, failures)
plotly_rga(fit_rga,
           main = "Crow-AMSAA Model",
           xlab = "Cumulative Time",
           ylab = "Cumulative Failures")
```

Hover over any data point to read the exact cumulative time and failure count. The fitted curve and confidence bounds can be toggled via the legend.

## Accelerated Life Testing Plots

These functions accept `alt.make()` and `alt.fit()` objects from `WeibullR.ALT` — see @sec-alt.

### ALT Probability Plot — `plotly_alt()`

```{r}
#| label: fig-alt
#| fig-cap: "Interactive ALT probability plot: Arrhenius-Lognormal model (Class B insulation)"
t41    <- NelsonData("table4.1")
set170 <- alt.data(t41$C170f, s = t41$C170s, stress = 170)
set190 <- alt.data(t41$C190f, s = t41$C190s, stress = 190)
set220 <- alt.data(t41$C220f, s = t41$C220s, stress = 220)

arrobj <- alt.make(list(set170, set190, set220),
                   dist = "lognormal", alt.model = "arrhenius",
                   method.fit = "mle", view_dist_fits = FALSE)
prlobj <- alt.parallel(arrobj)

plotly_alt(prlobj,
           main = "Arrhenius-Lognormal ALT Probability Plot",
           xlab = "Time to Failure (hours)",
           ylab = "Probability of Failure")
```

Each color represents a different temperature level. Hover to inspect unreliability at any time. Note: `plotly_alt()` requires a parallelized object from `alt.parallel()`.

### Life-Stress Relationship Plot — `plotly_rel()`

```{r}
#| label: fig-rel
#| fig-cap: "Interactive life-stress relationship plot"
lnrobj <- alt.fit(prlobj)
plotly_rel(lnrobj,
           main = "Life-Stress Relationship: Arrhenius Model",
           xlab = "Temperature (°C)",
           ylab = "Time to Failure")
```

Each line represents a percentile of the failure distribution (P10, P63.2, P90, etc.). Hover over any point to read the expected life at that stress level and percentile.

## Repairable Systems Plots

These functions accept the fitted objects from `ReliaGrowR` — the same objects produced in @sec-rs.

### NHPP Plot — `plotly_nhpp()`

```{r}
#| label: fig-nhpp
#| fig-cap: "Interactive Crow-AMSAA NHPP plot: industrial pump fleet"
times_pump    <- c(500, 1200, 2100, 3300, 4800, 6500, 8400, 10500, 12800, 15000)
failures_pump <- c(3, 4, 4, 5, 4, 3, 3, 3, 2, 2)

fit_nhpp <- nhpp(time = times_pump, event = failures_pump)
plotly_nhpp(fit_nhpp,
            main = "Crow-AMSAA: Industrial Pump Fleet",
            xlab = "Cumulative Operating Hours",
            ylab = "Cumulative Failures")
```

### MCF Plot — `plotly_mcf()`

```{r}
#| label: fig-mcf
#| fig-cap: "Interactive Mean Cumulative Function plot with confidence bounds"
pump_data <- data.frame(
  id    = c(1, 1, 1,  2, 2, 2,  3, 3,  4, 4, 4, 4,  5, 5),
  time  = c(310, 850, 1620,  420, 1050, 2100,  580, 1890,
            240, 710, 1380, 2400,  530, 1740),
  event = rep(1, 14)
)
end_times  <- c("1" = 3000, "2" = 3000, "3" = 3000, "4" = 3000, "5" = 3000)

mcf_result <- mcf(id = pump_data$id, time = pump_data$time,
                  event = pump_data$event, end_time = end_times)
plotly_mcf(mcf_result,
           main = "MCF: Industrial Pump Fleet",
           xlab = "Operating Hours",
           ylab = "Expected Cumulative Failures per System")
```

### Exposure Plot — `plotly_exposure()`

```{r}
#| label: fig-exposure
#| fig-cap: "Interactive exposure plot showing fleet-level event rate"
exp_result <- exposure(id = pump_data$id, time = pump_data$time,
                       event = pump_data$event)
plotly_exposure(exp_result,
                main = "Fleet Exposure Plot",
                xlab = "Operating Hours",
                ylab = "Event Rate (failures / hour)")
```

## Embedding in Reports and Dashboards

ReliaPlotR outputs are standard `plotly` objects and embed natively in:

**Quarto / R Markdown** — interactive widgets in HTML output, static fallback in PDF:

````markdown
```{{r}}
library(ReliaPlotR)
library(WeibullR)
failures <- c(30, 49, 82, 90, 96)
obj <- wblr.conf(wblr.fit(wblr(failures)))
plotly_wblr(obj)
```
````

**Shiny** — use `plotlyOutput()` / `renderPlotly()`:

```{r}
#| eval: false
library(shiny)
library(plotly)
library(ReliaPlotR)
library(WeibullR)

ui <- fluidPage(plotlyOutput("wblr_plot"))

server <- function(input, output) {
  output$wblr_plot <- renderPlotly({
    obj <- wblr.conf(wblr.fit(wblr(c(30, 49, 82, 90, 96))))
    plotly_wblr(obj)
  })
}

shinyApp(ui, server)
```

## When to Use ReliaPlotR

| Use ReliaPlotR when... | Use base WeibullR / ReliaGrowR plots when... |
|---|---|
| Delivering results to non-R stakeholders | Running quick exploratory analysis in a script |
| Embedding in an HTML report or dashboard | Generating PDF / print output directly |
| Stakeholders need to hover and inspect points | Batch-processing many analyses |
| Presenting in a web browser or Shiny app | Working in a non-HTML environment |

## Getting Help

```{r}
#| eval: false
help(package = "ReliaPlotR")
```

- CRAN page: <https://CRAN.R-project.org/package=ReliaPlotR>

## Summary

ReliaPlotR wraps every major fitted object in the ReliaLearnR ecosystem — `wblr`, `duane`, `rga`, `nhpp`, `mcf`, `exposure`, `alt.make`, `alt.fit` — in an interactive plotly chart. The pattern is consistent: fit your model with WeibullR, ReliaGrowR, or WeibullR.ALT as usual, then pass the result to the corresponding `plotly_*` function. All nine functions support the same customization arguments (titles, axis labels, colors, grid control) and return standard plotly objects compatible with Quarto, R Markdown, and Shiny.
