# AI-Assisted Reliability Analysis with MCP {#sec-mcp}

## Introduction

All of the analyses in this book require writing R code or navigating a Shiny app. **Model Context Protocol (MCP)** opens a third path: letting an AI assistant call your R functions directly during a conversation, so you can describe what you want in plain language and the AI runs the analysis, interprets the results, and generates the plots.

MCP is an open standard that defines how an AI assistant discovers and calls external tools in the same way a web browser calls REST APIs. Reliability workflows are a natural fit: fitting a growth model, predicting future failures, computing a fleet MCF, and saving an interactive plot are all discrete, well-defined operations that map cleanly to MCP tools.

This chapter shows how to expose `ReliaGrowR` [@ReliaGrowR] and `ReliaPlotR` [@ReliaPlotR] as an MCP server so that Claude Desktop, Claude Code, or any other MCP-compatible client can run reliability analyses on demand.

## Prerequisites

Install the MCP infrastructure packages. `ellmer` [@ellmer] provides the `tool()` constructor; `mcptools` [@mcptools] provides the server transport:

```{r}
#| eval: false
install.packages(c("ellmer", "mcptools"))
```

Ensure the reliability packages are also installed:

```{r}
#| eval: false
install.packages(c("ReliaGrowR", "ReliaPlotR", "htmlwidgets"))
```

## MCP Tools for ReliaGrowR

The `reliagrow_mcp_tools()` function below returns a named list of `ellmer::tool()` objects. Each tool is a thin wrapper: it accepts plain-text inputs (times and failures as comma-separated strings), calls the underlying `ReliaGrowR` function, extracts the key scalar results, and returns a plain list that the AI receives as JSON.

```{r}
#| eval: false
library(ellmer)
library(ReliaGrowR)

reliagrow_mcp_tools <- function() {

  parse_vec <- function(x) as.numeric(strsplit(trimws(x), "\\s*,\\s*")[[1]])

  list(

    fit_duane = tool(
      .name        = "fit_duane",
      .description = "Fit a Duane reliability growth model to cumulative failure data.
                      Returns the growth slope (beta), scale parameter (K), and
                      cumulative MTBF at the last observation.",
      times    = type_string("Comma-separated cumulative times (e.g. '100,200,300')"),
      failures = type_string("Comma-separated cumulative failure counts (e.g. '1,2,1')"),
      function(times, failures) {
        t <- parse_vec(times)
        f <- parse_vec(failures)
        fit <- duane(t, f)
        list(
          beta  = round(fit$beta, 4),
          K     = round(fit$K, 4),
          cmtbf = round(fit$K * max(t)^(fit$beta - 1), 2)
        )
      }
    ),

    fit_crow_amsaa = tool(
      .name        = "fit_crow_amsaa",
      .description = "Fit a Crow-AMSAA (NHPP Power Law) reliability growth model.
                      Returns beta (shape) and lambda (scale). Beta < 1 means
                      reliability is improving; beta > 1 means worsening.",
      times    = type_string("Comma-separated cumulative times"),
      failures = type_string("Comma-separated cumulative failure counts"),
      function(times, failures) {
        t <- parse_vec(times)
        f <- parse_vec(failures)
        fit <- rga(t, f, model_type = "Crow-AMSAA")
        list(
          beta   = round(fit$beta, 4),
          lambda = round(fit$lambda, 6)
        )
      }
    ),

    fit_piecewise_nhpp = tool(
      .name        = "fit_piecewise_nhpp",
      .description = "Fit a piecewise Crow-AMSAA model with an optional known breakpoint.
                      If no breakpoint is supplied, the change point is detected automatically.
                      Returns per-segment beta and lambda values.",
      times     = type_string("Comma-separated cumulative times"),
      failures  = type_string("Comma-separated cumulative failure counts"),
      breakpoint = type_string("Optional breakpoint time, or empty string for auto-detection"),
      function(times, failures, breakpoint = "") {
        t  <- parse_vec(times)
        f  <- parse_vec(failures)
        bp <- if (nchar(trimws(breakpoint)) == 0) NULL else as.numeric(breakpoint)
        fit <- if (is.null(bp)) {
          rga(t, f, model_type = "Piecewise NHPP")
        } else {
          rga(t, f, model_type = "Piecewise NHPP", breaks = bp)
        }
        list(
          segments   = length(fit$beta),
          beta       = round(fit$beta, 4),
          lambda     = round(fit$lambda, 6),
          breakpoints = fit$breaks
        )
      }
    ),

    predict_growth = tool(
      .name        = "predict_growth",
      .description = "Predict cumulative failures at future times from a fitted Crow-AMSAA model.",
      times      = type_string("Historical cumulative times used to fit the model"),
      failures   = type_string("Historical cumulative failure counts"),
      pred_times = type_string("Future times to predict at (comma-separated)"),
      function(times, failures, pred_times) {
        t  <- parse_vec(times)
        f  <- parse_vec(failures)
        pt <- parse_vec(pred_times)
        fit  <- rga(t, f)
        pred <- predict_rga(fit, times = pt)
        data.frame(
          time               = pt,
          predicted_failures = round(pred$predicted, 2),
          lower_95           = round(pred$lower, 2),
          upper_95           = round(pred$upper, 2)
        )
      }
    ),

    fit_nhpp = tool(
      .name        = "fit_nhpp",
      .description = "Fit a Crow-AMSAA Non-Homogeneous Poisson Process (NHPP) to
                      repairable systems fleet failure times. Returns beta and lambda.",
      failure_times = type_string("Comma-separated failure times across all systems (pooled)"),
      function(failure_times) {
        ft  <- parse_vec(failure_times)
        fit <- nhpp(time = ft, event = rep(1L, length(ft)))
        list(
          beta   = round(fit$beta, 4),
          lambda = round(fit$lambda, 6)
        )
      }
    ),

    compute_mcf = tool(
      .name        = "compute_mcf",
      .description = "Compute the non-parametric Mean Cumulative Function (MCF) for a fleet.
                      Provide system IDs, event times, event indicators (1=failure, 0=censored),
                      and end times. Returns the MCF table.",
      ids        = type_string("Comma-separated system IDs (e.g. '1,1,1,2,2,2')"),
      times      = type_string("Comma-separated event times"),
      events     = type_string("Comma-separated event indicators (1=failure, 0=censored)"),
      end_times  = type_string("Comma-separated end-of-observation times, one per unique system"),
      function(ids, times, events, end_times) {
        id  <- parse_vec(ids)
        t   <- parse_vec(times)
        ev  <- as.integer(parse_vec(events))
        et  <- parse_vec(end_times)
        uid <- sort(unique(id))
        end <- setNames(et, as.character(uid))
        result <- mcf(id = id, time = t, event = ev, end_time = end)
        df <- as.data.frame(result)
        df[, sapply(df, is.numeric)] <- round(df[, sapply(df, is.numeric)], 4)
        df
      }
    ),

    plan_rdt = tool(
      .name        = "plan_rdt",
      .description = "Plan a Reliability Demonstration Test (RDT). Calculates the required
                      test time to demonstrate a reliability target at a given confidence level.",
      target       = type_number("Target reliability (e.g. 0.90 for 90%)"),
      mission_time = type_number("Mission time in hours (or any consistent unit)"),
      conf_level   = type_number("Confidence level (e.g. 0.80 for 80%)"),
      beta         = type_number("Assumed Weibull shape parameter"),
      n            = type_integer("Number of test units"),
      function(target, mission_time, conf_level, beta, n) {
        plan <- rdt(
          target       = target,
          mission_time = mission_time,
          conf_level   = conf_level,
          beta         = beta,
          n            = n
        )
        list(
          required_test_time = round(plan$test_time, 1),
          target_reliability = target,
          mission_time       = mission_time,
          confidence_level   = conf_level,
          sample_size        = n
        )
      }
    )

  )
}
```

::: {.callout-note}
## What happens inside each tool

Each `tool()` call takes a `.name`, a `.description` that the AI reads to decide when to call the tool, named argument descriptors (`type_string()`, `type_number()`, etc.), and a handler function. The handler does the real work and returns a plain list or data frame — no raw R objects. The AI sees the returned value as structured JSON.
:::

## MCP Tools for ReliaPlotR

The `reliaplotr_mcp_tools()` function wraps the four main `ReliaGrowR`-backed plot functions. Because `plotly` widgets cannot be sent through a text protocol, each tool saves a self-contained HTML file and returns its path. The AI then reports the path to the user.

```{r}
#| eval: false
library(ellmer)
library(ReliaGrowR)
library(ReliaPlotR)
library(htmlwidgets)

reliaplotr_mcp_tools <- function() {

  parse_vec <- function(x) as.numeric(strsplit(trimws(x), "\\s*,\\s*")[[1]])

  save_plot <- function(widget, prefix) {
    outpath <- tempfile(pattern = prefix, fileext = ".html")
    htmlwidgets::saveWidget(widget, file = outpath, selfcontained = TRUE)
    list(path = outpath)
  }

  list(

    plot_duane = tool(
      .name        = "plot_duane",
      .description = "Fit a Duane model and save an interactive Duane plot as a
                      self-contained HTML file. Returns the file path.",
      times    = type_string("Comma-separated cumulative times"),
      failures = type_string("Comma-separated cumulative failure counts"),
      title    = type_string("Optional plot title"),
      function(times, failures, title = "Duane Reliability Growth") {
        t   <- parse_vec(times)
        f   <- parse_vec(failures)
        fit <- duane(t, f)
        p   <- plotly_duane(fit, main = title,
                            xlab = "Cumulative Time",
                            ylab = "Cumulative MTBF")
        save_plot(p, "duane_")
      }
    ),

    plot_crow_amsaa = tool(
      .name        = "plot_crow_amsaa",
      .description = "Fit a Crow-AMSAA model and save an interactive cumulative
                      failures plot as a self-contained HTML file. Returns the file path.",
      times    = type_string("Comma-separated cumulative times"),
      failures = type_string("Comma-separated cumulative failure counts"),
      title    = type_string("Optional plot title"),
      function(times, failures, title = "Crow-AMSAA Reliability Growth") {
        t   <- parse_vec(times)
        f   <- parse_vec(failures)
        fit <- rga(t, f)
        p   <- plotly_rga(fit, main = title,
                          xlab = "Cumulative Time",
                          ylab = "Cumulative Failures")
        save_plot(p, "rga_")
      }
    ),

    plot_nhpp = tool(
      .name        = "plot_nhpp",
      .description = "Fit a Crow-AMSAA NHPP to repairable system failure times and
                      save an interactive plot as a self-contained HTML file.",
      failure_times = type_string("Comma-separated pooled failure times"),
      title         = type_string("Optional plot title"),
      function(failure_times, title = "Crow-AMSAA NHPP") {
        ft  <- parse_vec(failure_times)
        fit <- nhpp(time = ft, event = rep(1L, length(ft)))
        p   <- plotly_nhpp(fit, main = title,
                           xlab = "Operating Hours",
                           ylab = "Cumulative Failures")
        save_plot(p, "nhpp_")
      }
    ),

    plot_mcf = tool(
      .name        = "plot_mcf",
      .description = "Compute the non-parametric MCF for a fleet and save an interactive
                      plot as a self-contained HTML file.",
      ids       = type_string("Comma-separated system IDs"),
      times     = type_string("Comma-separated event times"),
      events    = type_string("Comma-separated event indicators (1=failure, 0=censored)"),
      end_times = type_string("Comma-separated end-of-observation times, one per system"),
      title     = type_string("Optional plot title"),
      function(ids, times, events, end_times, title = "Mean Cumulative Function") {
        id  <- parse_vec(ids)
        t   <- parse_vec(times)
        ev  <- as.integer(parse_vec(events))
        et  <- parse_vec(end_times)
        uid <- sort(unique(id))
        end <- setNames(et, as.character(uid))
        result <- mcf(id = id, time = t, event = ev, end_time = end)
        p <- plotly_mcf(result, main = title,
                        xlab = "Operating Hours",
                        ylab = "Expected Cumulative Failures per System")
        save_plot(p, "mcf_")
      }
    )

  )
}
```

## Launching the MCP Server

Save the two tool-list functions and the launch call in a single script, for example `mcp-server.R`:

```{r}
#| eval: false
# mcp-server.R
library(ReliaGrowR)
library(ReliaPlotR)
library(ellmer)
library(mcptools)
library(htmlwidgets)

# --- paste reliagrow_mcp_tools() and reliaplotr_mcp_tools() here, or source() them ---

mcptools::mcp_server(
  tools = c(reliagrow_mcp_tools(), reliaplotr_mcp_tools())
)
```

Running this script starts an MCP server on the standard-IO transport, ready to accept tool calls from any connected client.

## Connecting to Claude Desktop

Claude Desktop reads its MCP server list from a JSON config file. On macOS, open or create `~/Library/Application Support/Claude/claude_desktop_config.json` and add an entry for the reliability server:

```json
{
  "mcpServers": {
    "reliagrow": {
      "command": "Rscript",
      "args": ["/absolute/path/to/mcp-server.R"]
    }
  }
}
```

Restart Claude Desktop. The ReliaGrowR and ReliaPlotR tools will appear in the tool palette and Claude will call them automatically when relevant.

## Connecting to Claude Code

From the terminal, register the server with Claude Code using the `claude mcp add` command:

```bash
claude mcp add -s user reliagrow -- Rscript /absolute/path/to/mcp-server.R
```

Verify the server is registered:

```bash
claude mcp list
```

From that point, any Claude Code session will have access to the reliability tools.

## Example AI Conversations

Once connected, you can have conversations like the following.

**Example 1 — Crow-AMSAA fit and interpretation:**

> **You:** I ran a reliability growth test. The cumulative times were 100, 200, 300, 400, 500 hours with 1, 2, 1, 3, 2 failures. Fit a Crow-AMSAA model and tell me whether reliability is improving.
>
> **Claude:** *(calls `fit_crow_amsaa`)* The fitted model gives **β = 0.87** and **λ = 0.023**. Because β < 1, cumulative failures are growing more slowly than linearly — reliability is **improving**. The system is responding well to corrective actions during test.

**Example 2 — Forecasting and test-continuation decision:**

> **You:** Using the same data, predict cumulative failures at 600, 800, and 1000 hours, and tell me whether I should continue testing.
>
> **Claude:** *(calls `predict_growth`)* At 600 hours: **10.2 failures** (95% CI: 8.1–12.7). At 800 hours: **12.9** (CI: 10.1–16.5). At 1000 hours: **15.4** (CI: 11.8–20.1). The growth trend is still active (β < 1). If your target is fewer than 15 cumulative failures at 1000 hours, you are on track — but the upper confidence bound exceeds 15, so additional test time would tighten that uncertainty.

**Example 3 — Interactive MCF plot for stakeholders:**

> **You:** I have a fleet of 3 pumps. IDs: 1,1,1,2,2,3,3. Times: 310,850,1620,420,1050,580,1890. All events are failures. End times: 3000,3000,3000. Generate an MCF plot I can share with the maintenance team.
>
> **Claude:** *(calls `plot_mcf`)* Done. The interactive MCF plot has been saved to `/var/folders/.../mcf_abc123.html`. Open that file in any web browser to explore it — hover over any point to see the exact expected cumulative failures per system. The MCF is concave upward across the fleet, indicating that failure rate is increasing with age; a preventive maintenance interval may be warranted.

## When to Use AI vs Scripted R

| Situation | AI + MCP | Scripted R |
|---|:---:|:---:|
| Quick one-off fit during a conversation | ✅ | |
| Explaining results to non-R stakeholders | ✅ | |
| Iterating on model choice interactively | ✅ | |
| Automating repeated batch analyses | | ✅ |
| Reproducible research / version control | | ✅ |
| Full custom plot formatting | | ✅ |
| Embedding results in a Quarto report | | ✅ |
| Large datasets or performance-sensitive work | | ✅ |

The AI + MCP path excels at conversational exploration, where you want to ask a question, get an answer in plain language, and then ask a follow-up. Scripted R remains the right choice for production pipelines and fully reproducible analyses.

## Getting Help

```{r}
#| eval: false
help(package = "ReliaGrowR")
help(package = "ReliaPlotR")
```

- `mcptools` GitHub: <https://github.com/posit-dev/mcptools>
- `ellmer` GitHub: <https://github.com/tidyverse/ellmer>
- MCP specification: <https://modelcontextprotocol.io>

## Summary

MCP lets an AI assistant call ReliaGrowR and ReliaPlotR functions directly during a conversation. The `reliagrow_mcp_tools()` function defines seven tools, Duane fit, Crow-AMSAA fit, piecewise NHPP, growth forecast, NHPP for repairable systems, MCF, and RDT planning,  that each return plain JSON results. The `reliaplotr_mcp_tools()` function adds four plot tools that save self-contained interactive HTML files. Register the server with Claude Desktop via `claude_desktop_config.json` or with Claude Code via `claude mcp add`, then describe your data in plain language and let the AI run the analysis.
