# Appendix: Helper Functions Reference {#sec-appendix .unnumbered}

The `ReliaLearnR` package exports five RAM calculation helper functions. They accept scalar or vector inputs, validate their arguments, and return a single numeric scalar.

```{r}
library(ReliaLearnR)
```

## rel() — Reliability

**Computes reliability as** $1 - \text{outage} / \text{total}$.

```{r}
#| eval: false
rel(outageTime, totalTime)
```

| Argument | Type | Description |
|---|---|---|
| `outageTime` | numeric scalar or vector | Forced outage (failed) time |
| `totalTime` | numeric scalar or vector | Total time in the same units |

**Returns**: Numeric scalar in $[0, 1]$. Sums vectors before dividing.

**Validation**: Errors on NA, non-numeric, negative values, zero `totalTime`, or mismatched vector lengths.

```{r}
# Motor: failed 10 days out of 5 years (1825 days)
rel(outageTime = 10, totalTime = 5 * 365)

# Vector input: two machines pooled together
# Returns (10+5)/(365+730) = 0.9863 — sums both machines before dividing
rel(outageTime = c(10, 5), totalTime = c(365, 730))
```

---

## avail() — Availability

**Computes availability as** $1 - \text{unavailable} / \text{total}$, where unavailable time includes both failures *and* scheduled maintenance.

```{r}
#| eval: false
avail(unavailTime, totalTime)
```

| Argument | Type | Description |
|---|---|---|
| `unavailTime` | numeric scalar or vector | Total unavailable time (failed + maintenance) |
| `totalTime` | numeric scalar or vector | Total time in the same units |

**Returns**: Numeric scalar in $[0, 1]$.

```{r}
# Motor: 10 days failed + 15 days maintenance over 5 years
avail(unavailTime = 10 + 15, totalTime = 5 * 365)

# Vector input: two assets
avail(unavailTime = c(5, 14), totalTime = c(3 * 365, 3 * 365))
```

---

## mttf() — Mean Time to Failure

**Computes MTTF as** $\text{total time} / \text{failures}$, for **non-repairable** items.

```{r}
#| eval: false
mttf(failures, totalTime)
```

| Argument | Type | Description |
|---|---|---|
| `failures` | numeric scalar or vector | Failure counts |
| `totalTime` | numeric scalar or vector | Total operating time |

**Returns**: Numeric scalar (same units as `totalTime`). Returns `Inf` with a warning if `failures = 0`.

```{r}
# 100 motors, 5 years each, 5 failures
mttf(failures = 5, totalTime = 100 * 5)

# Pooled fleet data
mttf(failures = c(2, 3), totalTime = c(500, 500))
```

---

## mtbf() — Mean Time Between Failures

**Computes MTBF as** $\text{total time} / \text{failures}$, for **repairable** systems.

The formula is identical to `mttf()` but the semantic interpretation differs — MTBF is the expected time *between* successive failures on the same repairable system.

```{r}
#| eval: false
mtbf(failures, totalTime)
```

| Argument | Type | Description |
|---|---|---|
| `failures` | numeric scalar or vector | Failure counts |
| `totalTime` | numeric scalar or vector | Total operating time |

**Returns**: Numeric scalar. Returns `Inf` with a warning if `failures = 0`.

```{r}
# Machine failed 5 times in 45,000 operating hours
mtbf(failures = 5, totalTime = 45000)

# Fleet of 50 generators, 8 years, 20 failures
mtbf(failures = 20, totalTime = 50 * 8)
```

---

## fr() — Failure Rate

**Computes failure rate (λ) as** $\text{failures} / \text{total time}$.

```{r}
#| eval: false
fr(failures, totalTime)
```

| Argument | Type | Description |
|---|---|---|
| `failures` | numeric scalar or vector | Failure counts |
| `totalTime` | numeric scalar or vector | Total operating time |

**Returns**: Numeric scalar (failures per unit time). Note: $\lambda = 1 / \text{MTBF}$.

```{r}
# 100 motors, 5000 hours each, 75 failures
fr(failures = 75, totalTime = 100 * 5000)

# Pooled data
fr(failures = c(10, 5), totalTime = c(1000, 2000))
```

---

## Validation Behaviour

All five functions share the same input checks via an internal `.validate_inputs()` helper:

| Condition | Error message |
|---|---|
| NA in either argument | `"NA values not allowed in inputs"` |
| Non-numeric argument | `'"argument" must be numeric'` |
| Negative values | `"inputs must be non-negative"` |
| Empty vector | `"inputs must not be empty vectors"` |
| Mismatched vector lengths | `"inputs must have the same length or one must be a scalar"` |
| Zero total time | `"sum(totalTime) is zero (cannot divide by zero)"` |

Each call below triggers a different validation rule and prints the corresponding error message inline:

```{r}
#| error: true
rel(NA, 100)      # NA in inputs
avail(-5, 100)    # negative value
mttf(5, 0)        # zero total time (division by zero)
```
