# Appendix: R Quick Reference {#sec-appendix .unnumbered}

This appendix lists every R package and function used in this book, organized by package, with the chapter(s) where each appears. Use it as a lookup reference when you want to find the right function quickly without having to search through the chapters.

Install all packages at once with:

```{r}
#| eval: false
install.packages(c(
  "ReliaLearnR",
  "WeibullR",
  "WeibullR.ALT",
  "ReliaGrowR",
  "DiagrammeR",
  "ReliaPlotR",
  "ReliaShiny"
))
```

---

## ReliaLearnR

RAM (Reliability, Availability, and Maintainability) calculation helpers. All five functions accept scalar or vector inputs and return a single numeric scalar.

```{r}
#| eval: false
install.packages("ReliaLearnR")
library(ReliaLearnR)
```

| Function | Description | Chapter |
|---|---|---|
| `rel(outageTime, totalTime)` | Reliability: $1 - \text{outage} / \text{total}$ | Ch. 1 |
| `avail(unavailTime, totalTime)` | Availability: $1 - \text{unavailable} / \text{total}$ (failures + maintenance) | Ch. 1 |
| `mttf(failures, totalTime)` | Mean Time to Failure for non-repairable items | Ch. 1 |
| `mtbf(failures, totalTime)` | Mean Time Between Failures for repairable systems | Ch. 1 |
| `fr(failures, totalTime)` | Failure rate $\lambda = \text{failures} / \text{total time}$ | Ch. 1 |

**Example usage:**

```{r}
library(ReliaLearnR)

rel(outageTime = 10, totalTime = 5 * 365)   # reliability over 5 years
avail(unavailTime = 25, totalTime = 5 * 365) # availability with maintenance
mttf(failures = 5, totalTime = 100 * 5)      # MTTF for a fleet of 100 units
mtbf(failures = 5, totalTime = 45000)        # MTBF for a repairable machine
fr(failures = 75, totalTime = 100 * 5000)    # failure rate (failures per hour)
```

---

## WeibullR

Life data analysis: Weibull probability plotting, distribution fitting, and confidence bounds.

```{r}
#| eval: false
install.packages("WeibullR")
library(WeibullR)
```

| Function | Description | Chapter |
|---|---|---|
| `wblr(...)` | Create a Weibull object from time/event data | Ch. 3, 7 |
| `wblr.fit(obj, ...)` | Fit a distribution (MLE or MRR) to a `wblr` object | Ch. 3, 7 |
| `wblr.conf(obj, ...)` | Compute confidence bounds on a fitted `wblr` object | Ch. 3, 7 |
| `MLEw2p(...)` | Fit a 2-parameter Weibull distribution via MLE directly | Ch. 3 |
| `plot_contour(obj)` | Static likelihood ratio contour plot | Ch. 3 |

**Example usage:**

```{r}
#| eval: false
library(WeibullR)

failures <- c(500, 1200, 900, 1500, 750)
da <- wblr(x = failures, s = c(2000, 2000))  # 2 suspensions
da <- wblr.fit(da, method.fit = "mle")
da <- wblr.conf(da, method.conf = "lrb")
plot(da)
```

---

## WeibullR.ALT

Accelerated Life Testing (ALT): fit life-stress relationships such as Arrhenius, Power Law, and Eyring to multi-stress datasets.

```{r}
#| eval: false
install.packages("WeibullR.ALT")
library(WeibullR.ALT)
```

| Function | Description | Chapter |
|---|---|---|
| `NelsonData(name)` | Load built-in Nelson ALT example datasets | Ch. 5, 7 |
| `alt.data(...)` | Prepare an ALT data object from raw time/event/stress data | Ch. 5, 7 |
| `alt.make(...)` | Create an ALT model object (specify distribution and relationship) | Ch. 5, 7 |
| `alt.parallel(obj)` | Constrain Weibull slopes to be equal across stress levels | Ch. 5, 7 |
| `alt.fit(obj, ...)` | Fit the life-stress relationship to the ALT model | Ch. 5, 7 |

**Example usage:**

```{r}
#| eval: false
library(WeibullR.ALT)

nd <- NelsonData("E2")
da <- alt.data(nd)
mo <- alt.make(da, distribution = "Weibull", relationship = "Arrhenius")
mo <- alt.parallel(mo)
mo <- alt.fit(mo)
plot(mo)
```

---

## ReliaGrowR

Reliability growth testing and repairable systems analysis: Crow-AMSAA, Duane, Power Law Process, and Mean Cumulative Function.

```{r}
#| eval: false
install.packages("ReliaGrowR")
library(ReliaGrowR)
```

| Function | Description | Chapter |
|---|---|---|
| `rdt(...)` | Plan a Reliability Demonstration Test | Ch. 4 |
| `rga(times, ...)` | Fit a Crow-AMSAA (NHPP Power Law) reliability growth model | Ch. 4, 7 |
| `predict_rga(obj, ...)` | Forecast cumulative failures from a `rga` object | Ch. 4 |
| `duane(times, ...)` | Fit a Duane reliability growth model | Ch. 4, 7 |
| `nhpp(...)` | Fit a Power Law Process to repairable systems data | Ch. 6, 7 |
| `mcf(...)` | Estimate the Mean Cumulative Function (Nelson-Aalen) | Ch. 6, 7 |
| `exposure(...)` | Compute fleet exposure from multi-system event data | Ch. 6, 7 |

**Example usage:**

```{r}
#| eval: false
library(ReliaGrowR)

# Reliability growth
times <- c(9.2, 25, 61.5, 260, 300, 710, 916, 1010, 1220)
rga_fit <- rga(times, T = 1500)
plot(rga_fit)

# Repairable systems
events <- data.frame(id = c(1,1,1,2,2), time = c(100,300,500,200,450))
mcf_fit <- mcf(events)
plot(mcf_fit)
```

---

## DiagrammeR

Graph and diagram visualisation: render Graphviz DOT diagrams for reliability block diagrams and fault trees.

```{r}
#| eval: false
install.packages("DiagrammeR")
library(DiagrammeR)
```

| Function | Description | Chapter |
|---|---|---|
| `grViz(diagram)` | Render a Graphviz DOT string as an SVG diagram | Ch. 2 |

**Example usage:**

```{r}
#| eval: false
library(DiagrammeR)

grViz("
  digraph series {
    rankdir = LR
    node [shape = rectangle]
    A -> B -> C
  }
")
```

---

## ReliaPlotR

Interactive, plotly-based versions of every static plot produced by WeibullR, WeibullR.ALT, and ReliaGrowR. Drop-in replacements that return `plotly` objects embeddable in Quarto documents and Shiny apps.

```{r}
#| eval: false
install.packages("ReliaPlotR")
library(ReliaPlotR)
```

| Function | Description | Chapter |
|---|---|---|
| `plotly_wblr(obj)` | Interactive Weibull probability plot | Ch. 7 |
| `plotly_contour(obj)` | Interactive likelihood ratio contour plot | Ch. 7 |
| `plotly_duane(obj)` | Interactive Duane reliability growth plot | Ch. 7 |
| `plotly_rga(obj)` | Interactive Crow-AMSAA plot | Ch. 7 |
| `plotly_alt(obj)` | Interactive ALT probability plot | Ch. 7 |
| `plotly_rel(obj)` | Interactive life-stress relationship plot | Ch. 7 |
| `plotly_nhpp(obj)` | Interactive NHPP intensity (recurrence rate) plot | Ch. 7 |
| `plotly_mcf(obj)` | Interactive Mean Cumulative Function plot | Ch. 7 |
| `plotly_exposure(obj)` | Interactive fleet exposure plot | Ch. 7 |

**Example usage:**

```{r}
#| eval: false
library(ReliaPlotR)
library(WeibullR)

da <- wblr(x = c(500, 1200, 900, 1500, 750))
da <- wblr.fit(da, method.fit = "mle")
da <- wblr.conf(da, method.conf = "lrb")
plotly_wblr(da)   # interactive version of plot(da)
```

---

## ReliaShiny

A point-and-click Shiny web application that wraps all of the above analyses in a browser-based UI — no coding required. Modules cover RAM, Life Data Analysis, Reliability Growth, ALT, Repairable Systems, and RBD.

```{r}
#| eval: false
install.packages("ReliaShiny")
library(ReliaShiny)
```

| Function | Description | Chapter |
|---|---|---|
| `ReliaShiny()` | Launch the ReliaShiny web application in the default browser | Ch. 8 |

**Example usage:**

```{r}
#| eval: false
library(ReliaShiny)
ReliaShiny()
```
